---
title: "Acute Kidney Injury (10x Visium) Application"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Acute Kidney Injury (10x Visium) Application}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  dpi = 72,
  dev = "png"
)
```

```{r setup}
library(STcompare)
library(SpatialExperiment)
library(MERINGUE)
library(rhdf5)
library(ggplot2)
library(dplyr)
library(patchwork)
library(Matrix)
library(BiocGenerics)
devtools::load_all()
```

# Introduction 

In the [`Getting Started with STcompare`](https://jef.works/STcompare/articles/getting-started-with-STcompare.html) tutorial, we demonstrated that `STcompare` works on both simulated and real ST data. We next sought to apply it to compare healthy and diseased datasets to discover potentially disease-relevant changes in spatial gene expression patterns. To this end, we focus on a common and serious disease, acute kidney injury (AKI), by comparing two previously published murine kidney sections assayed by spot-resolution 10x Visium: a section from a kidney 24 hours post initial AKI injury and a section from a sham surgery kidney as a control. 

This tutorial corresponds with Figures 2f-j from the [`STcompare paper`](https://doi.org/10.1101/2025.11.21.689847). 

The counts files are stored as `h5` files on Zenodo and can be downloaded directly into R from the `url` paths below.

# Load files

```{r load-data}

# The Visium mouse kidney datasets can be found Zenodo: https://doi.org/10.5281/zenodo.17676991
# IL3 - ischemia acute kidney injury dataset 
# NL3 - control dataset 

# AKI counts ###################################################################
zenodo_url <- "https://zenodo.org/records/19074288/files/IL3_filtered_feature_bc_matrix.h5?download=1"
temp_h5 <- tempfile(fileext = ".h5")
download.file(zenodo_url, temp_h5, mode = "wb")

# this is a list of the internal groups and datasets stored in the HDF5 file 
# we are interested in the "matrix/barcodes" dataset 
rhdf5::h5ls(temp_h5)

# Read the CSC-encoded matrix components from the HDF5 file
file_barcodes <- as.character(rhdf5::h5read(temp_h5, "matrix/barcodes"))
barcodes_matrix <- rhdf5::h5read(temp_h5, "matrix") # stores the sparse matrix components 

# convert CSC encoding into an R dgCMatrix
aki_counts <- Matrix::sparseMatrix(
  dims = barcodes_matrix$shape,
  i = as.numeric(barcodes_matrix$indices),
  p = as.numeric(barcodes_matrix$indptr),
  x = as.numeric(barcodes_matrix$data),
  index1 = FALSE
)
colnames(aki_counts) <- file_barcodes
rownames(aki_counts) <- barcodes_matrix[["features"]]$name

unlink(temp_h5)

# what the counts matrix looks like: 
aki_counts[1:5, 1:6]
```

```{r load-data2}
# Control counts ###############################################################

zenodo_url <- "https://zenodo.org/records/19074288/files/NL3_filtered_feature_bc_matrix.h5?download=1"
temp_h5 <- tempfile(fileext = ".h5")
download.file(zenodo_url, temp_h5, mode = "wb")

# Read the CSC-encoded matrix components from the HDF5 file
file_barcodes <- as.character(rhdf5::h5read(temp_h5, "matrix/barcodes"))
barcodes_matrix <- rhdf5::h5read(temp_h5, "matrix") # stores the sparse matrix components 

# convert CSC encoding into an R dgCMatrix
control_counts <- Matrix::sparseMatrix(
  dims = barcodes_matrix$shape,
  i = as.numeric(barcodes_matrix$indices),
  p = as.numeric(barcodes_matrix$indptr),
  x = as.numeric(barcodes_matrix$data),
  index1 = FALSE
)
colnames(control_counts) <- file_barcodes
rownames(control_counts) <- barcodes_matrix[["features"]]$name

unlink(temp_h5)
control_counts[1:5, 1:6]
```

The spatial positions are stored in `csv` files on Zenodo which can be downloaded directly into R from the `url` paths below. These text files contains a table with rows that correspond to spots. 

The columns correspond to the following fields:
  `barcode`: The sequence of the barcode associated to the spot
  `array_row`: The row coordinate of the spot range from 0 to 127 as the array has 128 rows.
  `array_col`: The column coordinate of the spot in the array. In order to express the orange crate arrangement of the spots, this column index uses even numbers from 0 to 222 for even rows, and odd numbers from 1 to 223 for odd rows with each row (even or odd) resulting in 111 spots.
  
```{r load-data3}
# positions ####################################################################
zenodo_url <- "https://zenodo.org/records/19074288/files/IL3_tissue_positions.csv?download=1"
aki_pos <- read.csv(zenodo_url, header = TRUE, stringsAsFactors = FALSE)

zenodo_url <- "https://zenodo.org/records/19074288/files/NL3_tissue_positions.csv?download=1"
ctrl_pos <- read.csv(zenodo_url, header = TRUE, stringsAsFactors = FALSE)

head(ctrl_pos)
head(aki_pos)
```

For this dataset in particular, this is an abridged version of the `tissue_positions.csv` provided by 10x Genomics. The spot rows have already been filtered to only include those in which the spot falls inside of the tissue. As such, we need to filter the counts matrices which is genes as rows x barcodes as columns to only include column names that are present in the barcodes of the positions matrices.

```{r load-data4}
# number of columns in counts is not the same number of rows in pos
dim(aki_counts)
dim(aki_pos)

dim(control_counts)
dim(ctrl_pos)

# limit the counts to the spots with positions and make sure they are in the same order
aki_counts <- aki_counts[, colnames(aki_counts) %in% aki_pos$barcode]
control_counts <- control_counts[, colnames(control_counts) %in% ctrl_pos$barcode]

# number of columns in counts is the same number of rows in pos
dim(aki_counts)
dim(aki_pos)

dim(control_counts)
dim(ctrl_pos)
```
To use functions that require that the barcodes are the row names of a matrix we will move the barcode column to be row names of the position matrices. Next, we visualize both control and AKI positions on the same plot. We rotate the positions by 90 degrees here simply to make the visualizations take up more space on the vertical axis than the horizontal axis.

```{r load-data5}
# plot positions ###############################################################

# set the barcode column as the rownames
rownames(ctrl_pos) <- ctrl_pos$barcode
ctrl_pos <- ctrl_pos[,-1]
colnames(ctrl_pos) <- c("x", "y")

rownames(aki_pos) <- aki_pos$barcode
aki_pos <- aki_pos[,-1]
colnames(aki_pos) <- c("x", "y")

ctrl_pos$group <- "Control"
aki_pos$group  <- "AKI"

# visualize both control and AKI positions on the same plot
df <- rbind(ctrl_pos, aki_pos)
pos_plot <- ggplot(df, aes(x = x, y = y, color = group)) +
  geom_point() +
  scale_color_manual(values = c("Control" = "blue", "AKI" = "red")) +
  coord_fixed() +
  labs(x = "x", y = "y", color = "Group") +
  theme_classic()
pos_plot

# rotate the tissue by 90-degrees for consistency with the paper 
ctrl_pos_rot <- transform(ctrl_pos, x = y, y = -x + max(ctrl_pos$x))
aki_pos_rot <- transform(aki_pos, x = y, y = -x + max(aki_pos$x))

df_rot <- rbind(ctrl_pos_rot, aki_pos_rot)
pos_rot_plot <- ggplot(df_rot, aes(x = x, y = y, color = group)) +
  geom_point(size=0.5) +
  scale_color_manual(values = c("Control" = "blue", "AKI" = "red")) +
  coord_fixed() +
  labs(x = "x", y = "y", color = "Group") +
  theme_classic()
pos_rot_plot
```
# Align

Since we want to eventually attribute spatial changes in gene expression patterns to AKI by comparing the same spatial locations in the control and AKI, it is important that we use an alignment method to make the matched spatial locations contain the same regions within the tissue and remove structural differences that could be confounding.  If we do not minimize structural misalignment, then we can less confidently conclude that the gene expression variation that we observe is due AKI.

Here, we used STalign to perform an affine-only alignment based on one-hot encoding of region labels: cortex, outer medulla, and inner medulla.
```{r, align}
# read the csv which has the AKI positions aligned to control via STalign
zenodo_url <- "https://zenodo.org/records/19486091/files/aki_region_onehot_STalign_to_ctrl_region_onehot_affine_only.csv.gz?download=1"
aki_STalign <- read.csv(gzcon(url(zenodo_url, open = "rb"), text = TRUE), header = TRUE, row.names =1)

# store the aligned positions
aki_pos_aligned <- aki_STalign[,c("aligned_x", "aligned_y")]

#rename columns and add group for consistency with ctrl_pos_rot
colnames(aki_pos_aligned) <- c("x", "y")
aki_pos_aligned$group  <- "AKI"
# rotate the tissue by 90-degrees for consistency with the paper 
aki_pos_rot <- transform(aki_pos_aligned, x = y, y = -x + max(aki_pos_aligned$x))
aki_pos_aligned <- aki_pos_rot

df_aligned <- rbind(ctrl_pos_rot, aki_pos_aligned)
pos_aligned_plot <- ggplot(df_aligned, aes(x = x, y = y, color = group)) +
  geom_point(size=0.5) +
  scale_color_manual(values = c("Control" = "blue", "AKI" = "red")) +
  coord_fixed() +
  labs(x = "x", y = "y", color = "Group") +
  theme_classic()
pos_aligned_plot
```

# Create SpatialExperiment 

Now that we have created counts matrices and aligned positions matrices we need to create a [`SpatialExperiment`](https://www.bioconductor.org/packages/release/bioc/html/SpatialExperiment.html) as this is the required input form for the functions we want to use in `STcompare` (and `SEraster`). A `SpatialExperiment` is an object that store spatial transcriptomics data, in this case, combining gene expression matrices as `assays` with the spatial coordinates of each spatial location as `spatialCoords`.

``` {r create_spatial_experiment1}
AKI_ctrl_SE <- SpatialExperiment::SpatialExperiment(
  assays = list(counts = control_counts),
  spatialCoords = as.matrix(ctrl_pos_rot[,1:2]),
)

AKI_aki_SE <- SpatialExperiment::SpatialExperiment(
  assays = list(counts = aki_counts),
  spatialCoords = as.matrix(aki_pos_aligned[,1:2])
)
```

# Rasterize
In order to use `STcompare` we need paired data to make one-to-one comparisons. The dimensions of the SpatialExperiment `assays` must be same for the control and AKI. And the row names of the `spatialCoords` for control must match the row names of the `spatialCoords` for AKI. And rows with the same name must have the same x,y coordinates.

So far we do not have true paired data because there is a different number of spots in control and AKI datasets and because the alignment caused the same barcode name in the control and AKI datasets no longer refer to the same x,y-coordinate.

Therefore, we need to create a new shared unit system that we can use to make one-to-one comparisons. 

In order to have units that we can use to make one-to-one comparisons, we need to aggregate the position and count matrices of control and AKI datasets to create a new share unit system in which each unit gets a unique identifier that refers to the same x,y coordinate in control and AKI. This aggregation process is also refered to as rasterization.

We will use `SEraster` to rasterize the data into a a new shared pixel grid, converting the matrices from spots-by-positions and genes-by-spots to pixel-by-positions and genes-by-pixels.

To use `SEraster::rasterizeGeneExpression`, the user should specify a resolution that results in pixel sizes that are relevant to the tissue size and the biological features of interest. The resolution should be larger enough to smooth out mismatches in the structural alignment but not too large where pixels are too big and thin biological structures of interest get lost.

`SEraster` also lets the user specify if using square or hexagonal pixels. Here we choose hexagonal pixels since they are closer in shape to the Visium spots.

Lastly, `SEraster::rasterizeGeneExpression` allows the user to specify what aggregation function they want to use. `fun` can be set to `mean` or `sum`. If `mean`, pixel value for each pixel would be mean of gene expression for all spots within the pixel. If `sum`, pixel value for each pixel would be sum of gene expression for all spots within the pixel.

Because we perform counts-per-million (CPM) normalization (converting the expression of any gene into a proportion of total gene expression for all genes in the pixel) after rasterization, we could have chosen either of `mean` or `sum`.

After performing CPM normalization, when we plot the total expression of all genes in each pixel, the value of each pixel should be 1 million (1e6). For `SEraster::plotRaster`, we specify the name of the assay from which we want to plot gene expression with `assay_name` and we specify the name to give the color bar with `name`. By default, if the the rasterized SpatialExperiment is not subsetted to one gene, `plotRaster` visualizes the total expression (from the specified assay) of all genes for each spot.

```{r, rasterize}
input <- list(AKI_ctrl = AKI_ctrl_SE, AKI_aki = AKI_aki_SE)
rast <- SEraster::rasterizeGeneExpression(
  input,
  resolution = 5,
  fun        = "sum",
  square     = FALSE,
  assay_name = 'counts'
)

# add a CPM normalization assay to the rasterized SpatialExperiment
assay(rast$AKI_ctrl, "CPM") <- Matrix::t(Matrix::t(assay(rast$AKI_ctrl))/Matrix::colSums(assay(rast$AKI_ctrl)))*1e6
assay(rast$AKI_aki, "CPM") <- Matrix::t(Matrix::t(assay(rast$AKI_aki))/Matrix::colSums(assay(rast$AKI_aki)))*1e6

# plot total expression for new pixels
p1 <- SEraster::plotRaster(rast$AKI_ctrl, assay_name = "CPM", name = "total expression")
p2 <- SEraster::plotRaster(rast$AKI_aki, assay_name = "CPM", name = "total expression")
p1 + p2
```

# Identify SVGs 

We now want to identify which genes are spatially variable genes (SVG). Existing SVG analysis methods can identify genes that change from spatially variable to not or vice versa across conditions. However, genes that are classified as spatially variable in both conditions can still vary in the specific spatial pattern in each condition. 

Unlike previous methods, `STcompare` is to be able to identify SVGs that change in spatial pattern across datasets.
To demonstrate this ability to identify that a gene's spatial pattern changed even if the gene is spatially variable in both datasets, we limit the analysis here to only genes that are spatially variable in both control and AKI. However, `STcompare` can be run without filtering for SVGs. 

We use Moran's I to identify SVGs and use the implementation from `MERINGUE`. Here we choose a `filterDist` size that is 2x the size of our rasterization resolution to achieve a neighbor-relationship graph that looks like the image below with connectivity to adjacent neighbors.

``` {r moransI}
# use MERINGUE to get the neighbor-relationships 
w <- MERINGUE::getSpatialNeighbors(SpatialExperiment::spatialCoords(rast$AKI_ctrl), filterDist = 10)
par(mfrow=c(1,1))
MERINGUE::plotNetwork(SpatialExperiment::spatialCoords(rast$AKI_ctrl), w)
```

Define a function to calculate Moran's I for a given SpatialExperiment (optionally specify assay name) and filter distance.
``` {r moransI1}
# Returns a dataframe with these columns: 
# observed: Observed Moran's I statistic measuring spatial autocorrelation
# expected: Expected Moran's I under null hypothesis of random distribution
# sd: Standard deviation of Moran's I under the null hypothesis
# p.value: Statistical significance of the spatial pattern
# p.adj: Adjusted p-value (FDR-corrected) for multiple testing correction
moransI <- function(SE_temp, filterDist, assayName=1){
  
  # use MERINGUE to get the neighbor-relationships 
  w <- MERINGUE::getSpatialNeighbors(SpatialExperiment::spatialCoords(SE_temp), filterDist = filterDist)
  par(mfrow=c(1,1))
  MERINGUE::plotNetwork(SpatialExperiment::spatialCoords(SE_temp), w)
  
  # Identify significantly spatially auto-correlated genes
  I <- MERINGUE::getSpatialPatterns(SummarizedExperiment::assays(SE_temp)[[assayName]], w)
  
  return(I)
}


```

Run Moran's I for control and AKI datasets separately to identify SVGs in each dataset. We specify the assay name as "CPM" to use the CPM normalized assay for this analysis.
``` {r moransI2}
moransI_ctrl <- moransI(rast$AKI_ctrl, filterDist = 10, assayName = "CPM")
head(moransI_ctrl)
```

``` {r moransI3}
moransI_aki <- moransI(rast$AKI_aki, filterDist = 10, assayName = "CPM")
head(moransI_aki)
```

Here we get the list of shared SVGs, those genes which are significantly spatially autocorrelated in both control and AKI
``` {r moransI4}
# Filter for the genes that are statistically significantly autocorrelated 
svg_ctrl <- moransI_ctrl %>% filter(p.adj == 0) %>% rownames()
svg_aki <- moransI_aki %>% filter(p.adj == 0) %>% rownames()
svg_int <- intersect(svg_ctrl, svg_aki) # genes that are SVGs in both conditions 
svg_uni <- union(svg_ctrl, svg_aki) # genes that are SVG in one condition but not the other 

# Out of 32285 genes: 
# there are 1673 genes that are SVGs in the control dataset 
length(svg_ctrl)

# there are 2318 genes that are SVGs in the aki dataset  
length(svg_aki)

# there are 1046 genes that are SVGs in both datasets   
length(svg_int)

# there are 2945 genes that are SVGs in either datasets 
length(svg_uni)
```

The functions we used in `MERINGUE` changed the gene names to be syntactically valid for R (prepending the character "X" if a name starts with number and replacing the "-" with "." ). Here we will return the names to the original form.
```{r}
#return names in `svg_int` that are not in `rownames(rast$AKI_ctrl)`
setdiff(svg_int, rownames(rast$AKI_ctrl))

rownames(rast$AKI_ctrl) <- make.names(rownames(rast$AKI_ctrl), unique = TRUE)
rownames(rast$AKI_aki)  <- make.names(rownames(rast$AKI_aki),  unique = TRUE)

#check if any differences remain
setdiff(svg_int, rownames(rast$AKI_ctrl))


```

# Filter lowly expression genes

In the case that you run `STcompare` without filtering for SVGs, you do need to apply some alternative filtering. STcompare's `SpatialCorrelation` does not work when there is zero expression of a gene expression as it cannot create permutations for the empirical p-value from zero expression so we must at least remove genes with zero expression. Furthermore, rarely-expressed genes that have noisy expression are not likely to be significantly similarly or differently spatially patterned across datasets. To reduce the computational time of creating permutations for these genes, we can for example use the code below to filter to only include genes that are detected in at least 5% of pixels in both samples in downstream analyses.

``` {r, create_spatial_experiment2}
# there are 32285 genes in both samples 
# which rows in the matrix that the gene is expressed in greater than 5% of all spots 
# this has to be true in both datasets to be considered a "good gene" 
good.genes <- names(which(
  (Matrix::rowSums(SummarizedExperiment::assay(rast$AKI_ctrl, 'CPM') > 0) / ncol(SummarizedExperiment::assay(rast$AKI_ctrl, 'CPM'))*100 > 5) &
  (Matrix::rowSums(SummarizedExperiment::assay(rast$AKI_aki, 'CPM') > 0) / ncol(SummarizedExperiment::assay(rast$AKI_aki, 'CPM'))*100 > 5)))
length(good.genes)
```

# Spatial Correlation 

Next we use STcompare to calculate Pearson's correlation coefficient for genes that were SVGs in both conditions to investigate whether there was a change in the spatial patterning of expression. 

`deltaX` and `deltaY` are the parameters that controlling the degree of smoothing in permutations of datasets X and Y, respectively. 
Delta is a proportion calculated by dividing k neighbors by N total observations in X, where k is the number of neighbors in the permutation of X that should be within the radius smoothed by the Gaussian kernel to achieve the amount ofautocorrelation present in the original X. 
If a single delta is not known, a sequence of deltas can be inputted and the best delta will be found such that it minimizes the sum of squares of the residuals between the variogram of the permutation generated from the delta and the variogram of the target.  By default, `STcompare` sets this sequence to `seq(0.1, 0.9, .1)` but here we add two smaller values to the sequence to have finer smoothing options for the permutations. 

`nPermutations` is the parameter that controls how many permutations to create for the empirical p-value estimation. The more permutations, the more accurate the p-value estimation but the longer the runtime. To balance this tradeoff, we can set `STcompare` to first create a smaller number of permutations and then only create a larger number of permutations for genes that have an empirical p-value less than 0.05 based on the smaller number of permutations. 

After computing the permutations for each gene, Multiple Hypothesis Testing (MHT) correction is used to correct for the false positives that arise from chance. The default MHT correction used is BH correction. To reduce the runtime for this tutorial, the results are precomputed.  

``` {r spatial-correlation1}
# set up parameters for spatialCorrelationGeneExp

# choosing the genes that are SVGs in both conditions 
genes_chosen <- svg_int

## alternative: choosing the genes that are expressed in greater than 5% of all spots
# genes_chosen <- good.genes

# The input is the spatial experiments subset to the chosen genes of interest
input <- list('AKI_ctrl'= rast$AKI_ctrl[genes_chosen,],
             'AKI_aki'= rast$AKI_aki[genes_chosen,])

```

``` {r spatial-correlation2, eval=FALSE}

# `deltaX` and `deltaY` are the parameters that controlling the degree of smoothing in permutations of datasets X and Y, respectively
deltaList <- replicate(length(genes_chosen), c(0.01, 0.05, seq(0.1, 0.9, .1)), simplify = FALSE)

# creates 100 permutations and then if a gene has an empirical p-value is less than 0.05, creates 1000 permutations for more accurate p-value estimation
nPermutations <- c(100,1000)

# running spatialCorrelationGeneExpIterPermutations 
start_time <- Sys.time()
deltaList <- replicate(length(genes_chosen), c(0.01, 0.05, seq(0.1, 0.9, .1)), simplify = FALSE)
kidneyCorrelation <- STcompare::spatialCorrelationGeneExpIterPermutations(
  input, 
  nPermutations = nPermutations, 
  deltaX = deltaList, 
  deltaY = deltaList, 
  assayName = "CPM", # use the CPM normalized assay for this analysis
  nThreads = 22, # parallelize genes across threads
  BPPARAM = BiocParallel::MulticoreParam(), #setting up parallel-execution back-end
  verbose = TRUE, # prints name of gene as progress updates, which is helpful for long runs
  seed = 0) # set seed for reproducibility of the permutation results
end_time <- Sys.time()
print(end_time - start_time) 

```

Now we can look at the results. 

`STcompare::spatialCorrelationGeneExp` returns two adjusted empirical p-values, `pValuePermuteX` from creating an empirical null from permutations
of observations in X and `pValuePermuteY` from creating an empirical null from permutations of observations in Y. We require both adjusted empirical p-values to be less than 0.05 to consider the gene statistically significantly correlated. 

The `correlationCoef` column contains the Pearson's correlation coefficient for the gene's spatial pattern across the two samples. A positive correlation coefficient indicates that the gene has a similar spatial pattern in both samples, while a negative correlation coefficient indicates that the gene has an opposite spatial pattern in the two samples. The magnitude of the correlation coefficient indicates how similar or different the spatial patterns are, with values closer to 1 or -1 indicating more similar or more different patterns, respectively.

We will save the names of the significantly positively correlated genes and significantly negatively correlated genes in separate vectors to use for downstream analyses. 

``` {r, spatial-correlation3}

# Full script to generate the precomputed vignette results available at:
# system.file("scripts", "visiumKidneySpatialCorrelation.R", package = "STcompare")

# load data generated from the script 
load(system.file("extdata", "kidneyCorrelation.RData", package = "STcompare"))

# look at results
head(kidneyCorrelation)

# saving names of the significantly positively correlated svg genes 
svgSigPos <- kidneyCorrelation %>%
  dplyr::filter(pValuePermuteX < 0.05 & pValuePermuteY < 0.05) %>%  
  dplyr::filter(correlationCoef > 0) %>% 
  rownames()
# 707 genes are SVGs that are significantly positively correlated 
length(svgSigPos)

# saving names of the significantly negatively correlated svg genes 
svgSigNeg <- kidneyCorrelation %>%
  dplyr::filter(pValuePermuteX < 0.05 & pValuePermuteY < 0.05) %>%  
  dplyr::filter(correlationCoef < 0) %>% 
  rownames()
# 24 genes are SVGs that are significantly negatively correlated 
length(svgSigNeg)
```

Here, we visualize the distribution of correlation coefficients for significantly positively correlated, significantly negatively correlated, and not significantly correlated genes.

```{r, spatial-correlation4}

# visualize the significantly positively correlate and significantly negatively correlated svg genes  
# Figure 2J - Rate of Significance for SVGs 
# add a column for significance category (SigPos, SigNeg, NotSig) based on the svgSigPos and svgSigNeg vectors
fig_2j <- kidneyCorrelation %>%
  dplyr::mutate(Sig = dplyr::case_when(rownames(kidneyCorrelation) %in% svgSigPos ~ "SigPos",
                                       rownames(kidneyCorrelation) %in% svgSigNeg ~ "SigNeg",
                                       .default = "NotSig")) %>%
  ggplot2::ggplot(ggplot2::aes(x= Sig, y = correlationCoef, color = Sig)) +
  ggplot2::geom_violin() +
  ggplot2::geom_jitter(width = 0.2, alpha = 0.1) +
  ggplot2::ylim(-1,1) +
  ggplot2::scale_color_manual(values = c("SigPos" = "green", "SigNeg" = "blue", "NotSig" = "grey")) +
  ggplot2::theme_classic() +
  ggplot2::labs(x = "Significance", 
                y = "Correlation Coefficient", 
                title = "Correlation Coefficient for Significantly Positively, Negatively, and Not Significantly Correlated SVGs")
fig_2j


```

# Spatial Fold Change Similarity 

Next, we use `STcompare::spatialSimilarity` to calculate the fold change similarity metric for the same genes that were SVGs in both conditions to investigate whether there is a magnitude change in the spatial patterning of expression.

```{r spatial-similarity}

# Finding the spatial similarity
# If t1 and t2 are null, then the default threshold is the 0.05 quantile of gene expression. 
start_time <- Sys.time()
ss <- STcompare::spatialSimilarity(
  input = input, 
  foldChange = 1, 
  assayName = "CPM", 
  t1 = NULL, 
  t2 = NULL
)
end_time <- Sys.time()
print(end_time-start_time)
```

We compare the correlation and similarity metrics for only the significantly correlated genes and showed that the correlation and similarity metrics offer orthogonal insights as both significantly positively and significantly negatively correlated genes can have a range of similarity values.

``` {r fig2k}

# 2k - Comparing Correlation Metric to Similiarty metric 

# create a df of correlation and similarity values for each gene 
sim_corr <- data.frame(gene = ss$similarityTable$gene[ss$similarityTable$gene %in% svg_int],
                  correlation = kidneyCorrelation$correlationCoef[rownames(kidneyCorrelation) %in% svg_int], 
                  similar = ss$similarityTable$percentSimilarity[ss$similarityTable$gene %in% svg_int])
head(sim_corr) 

sim_corr_plt <- sim_corr %>%
  dplyr::mutate(Sig = case_when(gene %in% svgSigPos ~ "SigPos",
                         gene %in% svgSigNeg ~ "SigNeg",
                         .default = "NotSig")) %>%
  dplyr::filter(Sig %in% c("SigPos", "SigNeg")) %>%
  ggplot2::ggplot() + 
  ggplot2::geom_point(ggplot2::aes(x = correlation, y = similar, color = Sig), alpha = 0.25, size = 1) +
  ggplot2::xlim(NA,1) +
  ggplot2::coord_fixed() +
  ggplot2::scale_color_manual(values = c("SigPos" = "green", "SigNeg" = "blue")) +
  ggplot2::theme_classic() + 
  ggplot2::labs(x = "Correlation" , y = "Similarity", 
                title = "Similarity for significantly positively and negatively correlated SVGs")

sim_corr_plt

```

# Visualization of Example Genes

Lastly, we visualize the correlation and similarity metrics for two example genes, 

1) one significantly positively correlated gene but with low similarity because expression in one condition is greater than the other for all pixels and
2) one significantly negatively correlated gene with the smallest difference in number of pixels for which expression in AKI is twice as much as control (y > 2x; `percentDissimilarityY`) and number of pixels for which expression in control is twice as much as AKI (x > 2y; `percentDissimilarityX`). 

```{r correlation-plots1}

compute_dissimilarity <- function(x, y) {
  data.frame(
    percentDissimilarityY = mean(y > 2 * x, na.rm = TRUE),
    percentDissimilarityX = mean(x > 2 * y, na.rm = TRUE)
  )
}

sim_corr_2 <- sim_corr %>%
  dplyr::rowwise() %>%
  dplyr::mutate(
    compute_dissimilarity(
      x = as.numeric(rast$ctrl[gene, ]),
      y = as.numeric(rast$AKI[gene, ])
    )
  ) %>%
  dplyr::ungroup()


# for the significantly positively correlated svg genes, get gene with lowest similarity (if tied break tie by highest correlation coefficient)
gene1 <- sim_corr_2 %>%
  dplyr::filter(gene %in% svgSigPos) %>% # filter to only significantly positively correlated genes
  dplyr::arrange(desc(correlation)) %>% # if there are ties in similarity, the gene with the highest correlation coefficient will be at the top
  dplyr::arrange(similar) %>% # arrange in ascending order of similarity so that the gene with the lowest similarity is at the top, 
  dplyr::pull(gene) %>%
  head(1)

# for the significantly negatively correlated svg genes, get gene with smallest difference in percent dissimilarity
gene2 <- sim_corr_2 %>%
  dplyr::filter(gene %in% svgSigNeg) %>% # filter to only significantly negatively correlated genes
  dplyr::mutate(diff = abs(percentDissimilarityX - percentDissimilarityY)) %>% # create a new column for the absolute difference between percentDissimilarityX and percentDissimilarityY
  dplyr::arrange(diff) %>% # arrange in ascending order of the difference column so that the gene with the smallest difference is at the top
  dplyr::pull(gene) %>%
  head(1)
```

Next, we use `SEraster::plotRaster` to plot gene expression for the two example genes in control and AKI samples to visualize the spatial patterns of expression. 
Then, we use `STcompare::plotCorrelationGeneExp` to visualize the correlation of the spatial patterns across the two samples for each gene. 


``` {r correlation-plots2}

# plot gene expression spatial patterns of example genes

#get the shared pixels between the two samples to show only the pixels that are used for the one-to-one comparisons
sharedPixels <- intersect(rownames(SpatialExperiment::spatialCoords(rast$AKI_ctrl)),
                           rownames(SpatialExperiment::spatialCoords(rast$AKI_aki)))

#use SEraster::plotRaster to plot the spatial pattern of gene expression for each gene in each sample for only the shared pixels
gene1_plt_ctrl  <- SEraster::plotRaster(rast$AKI_ctrl[gene1, sharedPixels], assay_name = "CPM", name = "CPM expression")
gene1_plt_aki <- SEraster::plotRaster(rast$AKI_aki[gene1, sharedPixels], assay_name = "CPM", name = "CPM expression")

gene2_plt_ctrl  <- SEraster::plotRaster(rast$AKI_ctrl[gene2, sharedPixels], assay_name = "CPM", name = "CPM expression")
gene2_plt_aki <- SEraster::plotRaster(rast$AKI_aki[gene2, sharedPixels], assay_name = "CPM", name = "CPM expression")


# use the plotCorrelationGeneExp function to generate the correlation plots 

gene1_corr <- plotCorrelationGeneExp(
  speList = input, 
  spatialCorrelation = kidneyCorrelation, 
  geneName = gene1, 
  assayName = "CPM"
  )
gene2_corr <- plotCorrelationGeneExp(
  speList = input, 
  spatialCorrelation = kidneyCorrelation, 
  geneName = gene2, 
  assayName = "CPM"
  )

# use patchwork to put three plots together for each gene
gene1_plt_ctrl | gene1_plt_aki | gene1_corr
gene2_plt_ctrl | gene2_plt_aki | gene2_corr
```

Lastly, we use `STcompare::linearRegression` to visualize the fold change relationship across pixels and classifications for eache pixel for each gene as a scatter plot and `STcompare::pixelClass` to show the spatial pattern of the pixel classification. In both cases, pixels are classified into categories based on whether they have higher expression in control, higher expression in AKI, or similar expression in both samples.

```{r correlation-plots3}
# Figure 2I - fold change linear regression and spot classification 
gene1_lr <- STcompare::linearRegression(ss, gene=gene1, assayName = "CPM") + ggplot2::theme_classic() + ggplot2::coord_fixed()
gene2_lr <- STcompare::linearRegression(ss, gene=gene2, assayName = "CPM") + ggplot2::theme_classic() + ggplot2::coord_fixed()

gene1_pc <- pixelClass(ss, gene=gene1, assayName = "CPM")
gene2_pc <- pixelClass(ss, gene=gene2, assayName = "CPM")

# use patchwork to put two plots together for each gene
gene1_lr | gene1_pc
gene2_lr | gene2_pc

```

















