---
title: "Acute Kidney Injury (10x Visium) Application"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{acute-kidney-injury-10x-visium}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 72,
  dev = "png",
  fig.path = "getting-started-with-STcompare-figures/"

)
```

```{r setup}
library(STcompare)
library(rhdf5)
library(Matrix)
library(ggplot2)
library(dplyr)

```

# Introduction 

In the [`Getting Started with STcompare`](https://jef.works/STcompare/articles/getting-started-with-STcompare.html) tutorial, having demonstrated that `STcompare` works on both simulated and real ST data in the, we next sought to apply it to compare healthy and diseased datasets to discover potentially disease-relevant changes in spatial gene expression patterns. To this end, we focus on a common and serious disease, acute kidney injury (AKI), by comparing two previously published murine kidney sections assayed by spot-resolution 10x Visium: a section from a kidney 24 hours post initial AKI injury and a section from a normal kidney as a control. 

This tutorial corresponds with Figures 2f-j from the [`STcompare paper`](https://doi.org/10.1101/2025.11.21.689847). 

The files are stored as a H5 file on Zenodo. We are assuming here that the counts and the position matrices are well formed and correspond properly with one another. 

```{r load-data}

# The Visium mouse kidney datasets can be found Zenodo: https://zenodo.org/records/17676992
# IL3 - ischemia acute kidney injury dataset 
# NL3 - control dataset 

# AKI counts ###################################################################
zenodo_url <- "https://zenodo.org/records/17676992/files/IL3_filtered_feature_bc_matrix.h5?download=1"
temp_h5 <- tempfile(fileext = ".h5")
download.file(zenodo_url, temp_h5, mode = "wb")

# this is a list of the internal groups and datasets stored in the HDF5 file 
# we are interested in the "matrix/barcodes" dataset 
rhdf5::h5ls(temp_h5)

# Read the CSC-encoded matrix components from the HDF5 file
file_barcodes <- as.character(rhdf5::h5read(temp_h5, "matrix/barcodes"))
barcodes_matrix <- rhdf5::h5read(temp_h5, "matrix") # stores the sparse matrix components 

# convert CSC encoding into an R dgCMatrix
aki_counts <- Matrix::sparseMatrix(
  dims = barcodes_matrix$shape,
  i = as.numeric(barcodes_matrix$indices),
  p = as.numeric(barcodes_matrix$indptr),
  x = as.numeric(barcodes_matrix$data),
  index1 = FALSE
)

colnames(aki_counts) <- file_barcodes
rownames(aki_counts) <- as.data.frame(barcodes_matrix[["features"]])$name

unlink(temp_h5)


# Control counts ###############################################################

zenodo_url <- "https://zenodo.org/records/17676992/files/NL3_filtered_feature_bc_matrix.h5?download=1"
temp_h5 <- tempfile(fileext = ".h5")
download.file(zenodo_url, temp_h5, mode = "wb")

# Read the CSC-encoded matrix components from the HDF5 file
file_barcodes <- as.character(rhdf5::h5read(temp_h5, "matrix/barcodes"))
barcodes_matrix <- rhdf5::h5read(temp_h5, "matrix") # stores the sparse matrix components 

# convert CSC encoding into an R dgCMatrix
control_counts <- Matrix::sparseMatrix(
  dims = barcodes_matrix$shape,
  i = as.numeric(barcodes_matrix$indices),
  p = as.numeric(barcodes_matrix$indptr),
  x = as.numeric(barcodes_matrix$data),
  index1 = FALSE
)

colnames(control_counts) <- file_barcodes
rownames(control_counts) <- as.data.frame(barcodes_matrix[["features"]])$name

unlink(temp_h5)

head(aki_counts)
head(control_counts)

# positions ####################################################################

zenodo_url <- "https://zenodo.org/records/17676992/files/IL3_tissue_positions.csv?download=1"
aki_pos <- read.csv(zenodo_url, header = TRUE, stringsAsFactors = FALSE)

zenodo_url <- "https://zenodo.org/records/17676992/files/NL3_tissue_positions.csv?download=1"
ctrl_pos <- read.csv(zenodo_url, header = TRUE, stringsAsFactors = FALSE)

# given the counts the the position matrix 
head(colnames(aki_counts))
head(aki_pos$barcode)

# Standardize spot identifiers by removing sample-specific prefixes
# The regex "^[^_]*_" matches from the start (^) through the first underscore:
#   [^_]* = any characters except underscore
#   _     = the first underscore itself
# Original format: "SAMPLE_X_Y" 
# Standardized format: "X_Y"
aki_pos$barcode <- sub("^[^_]*_", "", aki_pos$barcode)
ctrl_pos$barcode <- sub("^[^_]*_", "", ctrl_pos$barcode)

# limit the counts to the spots with positions and make sure they are in the same order

# AKI 
# aki_pos$barcode with tag removed and colnames(aki_counts) are not identical 
identical(aki_pos$barcode, colnames(aki_counts))

# There are spots in aki_counts that are not in aki_pos
setdiff(aki_pos$barcode, colnames(aki_counts))
setdiff(colnames(aki_counts), aki_pos$barcode)

# removing the spots that are in aki_counts but not in aki_pos 
# remove the tag from the position barcodes 
aki_counts <- aki_counts[, colnames(aki_counts) %in% aki_pos$barcode]

# they are now identical, if this was not the case, more alignments 
# step would've been needed 
identical(sub("^[^_]+_", "", aki_pos$barcode), colnames(aki_counts))

# Control 
# The control is also not identical 
identical(ctrl_pos$barcode, colnames(control_counts))

# same case as before, remove the spots that are in the counts, but no in the pos
setdiff(ctrl_pos$barcode, colnames(control_counts))
setdiff(colnames(control_counts), ctrl_pos$barcode)
control_counts <- control_counts[, colnames(control_counts) %in% ctrl_pos$barcode]

# they are now identical
identical(ctrl_pos$barcode, colnames(control_counts))

# plot positions ###############################################################

# set the barcode column as the rownames
rownames(ctrl_pos) <- ctrl_pos$barcode
ctrl_pos <- ctrl_pos[,-1]
colnames(ctrl_pos) <- c("x", "y")

rownames(aki_pos) <- aki_pos$barcode
aki_pos <- aki_pos[,-1]
colnames(aki_pos) <- c("x", "y")

ctrl_pos$group <- "Control"
aki_pos$group  <- "AKI"

df <- rbind(ctrl_pos, aki_pos)
pos_plot <- ggplot(df, aes(x = x, y = y, color = group)) +
  geom_point() +
  scale_color_manual(values = c("Control" = "blue", "AKI" = "red")) +
  coord_fixed() +
  labs(x = "x", y = "y", color = "Group") +
  theme_classic()
pos_plot


# rotate the tissue by 90-degrees for consistency with the paper 

ctrl_pos_rot <- transform(ctrl_pos, x = y, y = -x + max(ctrl_pos$x))
aki_pos_rot <- transform(aki_pos, x = y, y = -x + max(aki_pos$x))

df_rot <- rbind(ctrl_pos_rot, aki_pos_rot)
pos_rot_plot <- ggplot(df_rot, aes(x = x, y = y, color = group)) +
  geom_point(size=0.5) +
  scale_color_manual(values = c("Control" = "blue", "AKI" = "red")) +
  coord_fixed() +
  labs(x = "x", y = "y", color = "Group") +
  theme_classic()
pos_rot_plot

```

Now that it has been verified that both the counts and the position matrices are well formed, the inputs to `STcompare` is a [`SpatialExperiment`]("https://www.bioconductor.org/packages/release/bioc/html/SpatialExperiment.html"). They are objects that store spatial transcriptomics data, in this case, combining gene expression matrices with the spatial coordinates of each spatial location. 

``` {r create_spatial_experiment}

# Turn the position and gene expression counts matrix into a spatial experiment 
# with the following assays 
#   counts: raw gene expression counts
#   libnorm: CPM (counts per million) normalization to adjust for sequencing depth

# remove the group column from position 
ctrl_pos_rot <- ctrl_pos_rot[, -3]
aki_pos_rot <- aki_pos_rot[, -3]

AKI_ctrl_SE <- SpatialExperiment::SpatialExperiment(
  assays = list(counts = control_counts,
                libnorm = Matrix::t(Matrix::t(control_counts)/Matrix::colSums(control_counts)*1e6)),
  spatialCoords = as.matrix(ctrl_pos_rot),
  
)

AKI_irl_SE <- SpatialExperiment::SpatialExperiment(
  assays = list(counts = aki_counts,
                libnorm = Matrix::t(Matrix::t(aki_counts)/Matrix::colSums(aki_counts))*1e6),
  spatialCoords = as.matrix(aki_pos_rot)

)

```

Removes rarely-expressed genes that likely represent noise by filtering to genes that are detected in at least 1% of spots in both samples.

``` {r}

# there are 32285 genes in both samples 
dim(SummarizedExperiment::assays(AKI_ctrl_SE)$libnorm) 
dim(SummarizedExperiment::assays(AKI_irl_SE)$libnorm) 

# which rows in the matrix that (1) has gene expression and (2) the gene is expressed in greater than 1% of all spots 
# this has to be true in both datasets to be considered a "good gene" 
good.genes <- names(which((
  Matrix::rowSums(SummarizedExperiment::assays(AKI_ctrl_SE)$libnorm > 0)) / dim(AKI_ctrl_SE)[2]*100 > 1 &
  (Matrix::rowSums(SummarizedExperiment::assays(AKI_irl_SE)$libnorm > 0))/dim(AKI_irl_SE)[2]*100 > 1))
length(good.genes)

# Reduced the number of genes from 32285 --> 12697 genes 

```

Of the "good genes" that have at least 1%, we now want to identify which of those genes are then spatially variable (SVG). We use Moran's I to identify SVGs. 

Unlike previous methods, `STcompare` is to be able to identify SVGs that change in spatial pattern across datasets. 

``` {r}

moransI <- function(SE_temp){
  
  # use MERINGUE to get the neighbor-relationships 
  w <- MERINGUE::getSpatialNeighbors(SpatialExperiment::spatialCoords(SE_temp), filterDist = 1.75)
  par(mfrow=c(1,1))
  MERINGUE::plotNetwork(SpatialExperiment::spatialCoords(SE_temp), w)
  
  # Identify significantly spatially auto-correlated genes
  start_time <- Sys.time()
  I <- MERINGUE::getSpatialPatterns(SummarizedExperiment::assays(SE_temp)$libnorm, w)
  end_time <- Sys.time()
  print(end_time - start_time)
  
  results.filter <- MERINGUE::filterSpatialPatterns(mat = SummarizedExperiment::assays(SE_temp)$libnorm,
                                          I = I,
                                          w = w,
                                          adjustPv = TRUE,
                                          alpha = 0.05,
                                          minPercentCells = 0.01,
                                          verbose = TRUE, details = TRUE)

  return(results.filter)
}


start_time <- Sys.time()
moransI_ctrl <- moransI(AKI_ctrl_SE[good.genes,])
end_time <- Sys.time()
print(end_time - start_time) 

start_time <- Sys.time()
moransI_irl <- moransI(AKI_irl_SE[good.genes,])
end_time <- Sys.time()
print(end_time - start_time) 

# "moransI_ctrl" and "moransI_irl" are dataframes these columns: 
# observed: Observed Moran's I statistic measuring spatial autocorrelation
#           Positive values = spatially clustered expression pattern
#           Values range from -1 to 1

# expected: Expected Moran's I under null hypothesis of random distribution
#           Typically close to zero

# sd: Standard deviation of Moran's I under the null hypothesis
#     Used to calculate the test statistic

# p.value: Statistical significance of the spatial pattern
#          Lower values = stronger evidence of non-random spatial expression

# p.adj: Adjusted p-value (FDR-corrected) for multiple testing correction

# minPercentCells: Minimum percentage of cells/spots where gene is expressed

head(moransI_ctrl)
head(moransI_irl)

# Running Moran's I already filtered the dataframe to only the genes that are statistically significantly autocorrelated 
# SVGs are then chosen as the gene that have expression patterns that possess more than 1% of all spots  
svg_ctrl <- moransI_ctrl %>% filter(minPercentCells > 0.01) %>% rownames()
svg_irl <- moransI_irl %>% filter(minPercentCells > 0.01) %>% rownames()
svg_int <- intersect(svg_ctrl, svg_irl) # genes that are SVGs in both conditions 
svg_uni <- union(svg_ctrl, svg_irl) # genes that are SVG in one condition but not the other 

# Out of 32285 gene: 
# there are 3389 genes that are SVGs in the control dataset 
length(svg_ctrl)

# there are 4447 genes that are SVGs in the aki dataset  
length(svg_irl)

# there are 2278 genes that are SVGs in both datasets   
length(svg_int)

# there are 5558 genes that are SVGs in either datasets 
length(svg_uni)

```

``` {r remove}
wd <- "/Users/vivien/Library/CloudStorage/OneDrive-JohnsHopkins/JefWorks/ST_compare"

# TODO: remove 
# loading in the moran's I data 

# saved as "moransI_ctrl" "moransI_irl" 
load(file.path(wd,"data/kidney_data/moransI_svg01per_20250817.RData"))

```


With the SVGs, use STcompare to calculate Pearson's correlation coefficient for genes that were SVGs in both conditions to investigate whether there was a change in the spatial patterning of expression. 

``` {r}

# TODO: did not run this for the sake of time 

# choosing the gene that are SVGs in both conditions 
genes_chosen <- svg_int

# The input is the spatial experiment for the genes that are SVGs in both conditions 
input <- list('AKI_ctrl'= AKI_ctrl_SE[genes_chosen, ], 
             'AKI_irl'= AKI_irl_SE[genes_chosen, ])

set.seed(0)
start_time <- Sys.time()
deltaList <- replicate(length(rownames(input[[1]])), c(0.01, 0.05, seq(0.1, 0.9, .1)), simplify = FALSE)
kidneyCorrelation <- STcompare::spatialCorrelationGeneExp(
  input, 
  nPermutations = 1000, # increased the number of permutations to account for stochasticity 
  deltaX = deltaList, 
  deltaY = deltaList, 
  returnPermutations = FALSE, 
  assayName = "libnorm", 
  nThreads = 5
)
end_time <- Sys.time()
print(end_time - start_time) 

# TODO: add change number of threads 



# TODO: remove 
# load kidneyCorrelation variable 
load(file.path(wd,"data/kidney_data/kidneyCorrelation_20250816.RData"))
head(kidneyCorrelation)

# 527 genes are SVGs that are significantly positively correlated 
kidneyCorrelation %>%
  dplyr::filter(rownames(kidneyCorrelation) %in% svg_int) %>%
  dplyr::filter(pValuePermuteX < 0.05 & pValuePermuteY < 0.05) %>%
  dplyr::filter(correlationCoef > 0) %>% 
  dim()

svgSigPos <- kidneyCorrelation %>%
  dplyr::filter(rownames(kidneyCorrelation) %in% svg_int) %>%
  dplyr::filter(pValuePermuteX < 0.05 & pValuePermuteY < 0.05) %>%  
  dplyr::filter(correlationCoef > 0) %>% 
  rownames()


# 25 genes are SVGs that are significantly negatively correlated 
kidneyCorrelation %>%
  dplyr::filter(rownames(kidneyCorrelation) %in% svg_int) %>%
  dplyr::filter(pValuePermuteX < 0.05 & pValuePermuteY < 0.05) %>%  
  dplyr::filter(correlationCoef < 0) %>% 
  dim()

svgSigNeg <- kidneyCorrelation %>%
  dplyr::filter(rownames(kidneyCorrelation) %in% svg_int) %>%
  dplyr::filter(pValuePermuteX < 0.05 & pValuePermuteY < 0.05) %>%  
  dplyr::filter(correlationCoef < 0) %>% 
  rownames()


# TODO: add comments 
# Figure 2J 
fig_2j <- kidneyCorrelation %>%
  dplyr::mutate(Sig = dplyr::case_when(rownames(kidneyCorrelation) %in% svgSigPos ~ "SigPos",
                         rownames(kidneyCorrelation) %in% svgSigNeg ~ "SigNeg",
                         .default = "")) %>%
  dplyr::mutate(pValueEmpirical = dplyr::case_when(pValuePermuteY > pValuePermuteX ~ pValuePermuteY,
                                     .default = pValuePermuteX)) %>%
  dplyr::mutate(pValueEmpiricalRound = dplyr::case_when(pValueEmpirical == 0 ~ 0.01,
                                    .default = pValueEmpirical)) %>%
  ggplot2::ggplot() + 
  ggplot2::geom_point(ggplot2::aes(x= correlationCoef, y = -log10(pValueEmpiricalRound), color = Sig), alpha = 0.25, size = 1) +
  ggplot2::xlim(NA,1) +
  ggplot2::scale_color_manual(values = c("SigPos" = "green", "SigNeg" = "blue")) +
  ggplot2::theme_classic() +
  ggplot2::geom_hline(yintercept = -log10(0.05), linetype = 'dashed', color = "black")  + 
  ggplot2::labs(x = "Correlation" , y = "-log10(p-value)", 
  )
fig_2j


# Adding a column for the empirical p-value (pValueEmpirical) 
# defined as the larger of the two x and y permuted p-values (greater of pValuePermuteX and pValuePermuteY)
kidneyCorrelationEmpirical <- kidneyCorrelation %>%
  dplyr::mutate(pValueEmpirical = dplyr::case_when(pValuePermuteY > pValuePermuteX ~ pValuePermuteY,
                                     .default = pValuePermuteX))


```

``` {r}

# Spots that are shared between the two datasets 
sharedPixels <- intersect(rownames(SpatialExperiment::spatialCoords(input[[1]])),
                          rownames(SpatialExperiment::spatialCoords(input[[2]])))

gene <- "Acsm2"

g1 <- SummarizedExperiment::assay(input[[1]], "libnorm")[gene, sharedPixels]
g2 <- SummarizedExperiment::assay(input[[2]], "libnorm")[gene, sharedPixels]

df12 <- cbind(g1, g2)

cor1 <- ggplot2::ggplot(df12, ggplot2::aes(x=g1, y=g2)) +
  ggplot2::geom_point(alpha = 0.2, size = 1) + ggplot2::theme_classic() + 
  #coord_fixed() +
  ggplot2::labs(x = paste0("Expression of pixel in ", names(input)[1]) , y = paste0("Expression of pixel in ", names(input)[2]),
       title = paste(" r = ", round(kidneyCorrelationEmpirical[gene,"correlationCoef"], 3),
                     " naĂŻve p = ", round(kidneyCorrelationEmpirical[gene,"pValueNaive"], 3),
                     " p = ", round(kidneyCorrelationEmpirical[gene, "pValueEmpirical"], 3))
  ) 

cor1

``` 

``` {r}

gene <- "Cbr1"

g1 <- SummarizedExperiment::assay(input[[1]], "libnorm")[gene, sharedPixels]
g2 <- SummarizedExperiment::assay(input[[2]], "libnorm")[gene, sharedPixels]

df12 <- cbind(g1, g2)

cor2 <- ggplot2::ggplot(df12, ggplot2::aes(x=g1, y=g2)) +
  ggplot2::geom_point(alpha = 0.2, size = 1) + ggplot2::theme_classic() + 
  ggplot2::labs(x = paste0("Expression of pixel in ", names(input)[1]) , y = paste0("Expression of pixel in ", names(input)[2]),
       title = paste(" r = ", round(kidneyCorrelationEmpirical[gene,"correlationCoef"], 3),
                     " naĂŻve p = ", round(kidneyCorrelationEmpirical[gene,"pValueNaive"], 3),
                     " p = ", round(kidneyCorrelationEmpirical[gene, "pValueEmpirical"], 3))
  ) 

cor2


```

``` {r}
# 2h ####################################################################


# Function to plot gene correlation between two samples
plot_gene_correlation <- function(gene, input, sharedPixels, kidneyCorrelationEmpirical) {
  #  expression data for the gene from both samples
  g1 <- SummarizedExperiment::assay(input[[1]], "libnorm")[gene, sharedPixels]
  g2 <- SummarizedExperiment::assay(input[[2]], "libnorm")[gene, sharedPixels]
  df12 <- cbind(g1, g2)
  
  # correlation plot
  cor_plot <- ggplot2::ggplot(df12, ggplot2::aes(x = g1, y = g2)) +
    ggplot2::geom_point(alpha = 0.2, size = 1) + 
    ggplot2::theme_classic() + 
    ggplot2::labs(
      x = paste0("Expression of pixel in ", names(input)[1]), 
      y = paste0("Expression of pixel in ", names(input)[2]),
      title = paste(" r = ", round(kidneyCorrelationEmpirical[gene, "correlationCoef"], 3),
                    " naĂŻve p = ", round(kidneyCorrelationEmpirical[gene, "pValueNaive"], 3),
                    " p = ", round(kidneyCorrelationEmpirical[gene, "pValueEmpirical"], 3))
    )
  
  return(cor_plot)
}

cor1 <- plot_gene_correlation("Acsm2", input, sharedPixels, kidneyCorrelationEmpirical)
cor2 <- plot_gene_correlation("Cbr1", input, sharedPixels, kidneyCorrelationEmpirical)

cor1
cor2

```

Cbr1 
- negative correlation 
- low fold change similarity 

Acsm2 
- positive correlation 
- low fold change similarity 



``` {r}

# Finding the spatial similarity

# TODO: t1 and t2, default 
start_time <- Sys.time()
ss <- STcompare::spatialSimilarity(input, foldChange = 1, assayName = "libnorm", t1 = NULL, t2 = NULL)
end_time <- Sys.time()
print(end_time-start_time)

# 2i ####################################################
lr1 <- STcompare::linearRegression(ss, gene="Acsm2", assayName = "libnorm") + ggplot2::theme_classic() + ggplot2::coord_fixed()
lr1 

lr2 <- STcompare::linearRegression(ss, gene="Cbr1", assayName = "libnorm") + ggplot2::theme_classic() + ggplot2::coord_fixed()
lr2 

pc1 <- pixelClass(ss, gene="Acsm2", assayName = "libnorm")
pc1

pc2 <- pixelClass(ss, gene="Cbr1", assayName = "libnorm")
pc2


# 2g ####################################################

df3 <- data.frame(SpatialExperiment::spatialCoords(AKI_ctrl_SE)[sharedPixels,],
                  dataset = "ctrl_aligned", color = SummarizedExperiment::assay(input[[1]], "libnorm")[gene, sharedPixels])
df4 <- data.frame(SpatialExperiment::spatialCoords(AKI_irl_SE)[sharedPixels,],
                  dataset = "irl_aligned", color = SummarizedExperiment::assay(input[[2]], "libnorm")[gene, sharedPixels])

pctrl_gene <- ggplot2::ggplot(df3, ggplot2::aes(x = x, y = y, color = color)) +
  ggplot2::geom_point(size = 0.7, alpha = 1, shape = 18) +
  viridis::scale_color_viridis() +
  ggplot2::coord_fixed() +
  ggplot2::theme_void()

pctrl_gene

pirl_gene <- ggplot2::ggplot(df4, ggplot2::aes(x = x, y = y, color = color)) +
  ggplot2::geom_point(size = 0.7, alpha = 1, shape = 18) +
  viridis::scale_color_viridis() +
  ggplot2::coord_fixed() +
  ggplot2::theme_void()

pirl_gene

# 2k #################################################################
# TODO: add gridExtra 

df5 <- data.frame(gene = ss$similarityTable$gene[ss$similarityTable$gene %in% svg_int],
                  correlation = kidneyCorrelation[svg_int, "correlationCoef" ], 
                  similar = ss$similarityTable$percentSimilarity[ss$similarityTable$gene %in% svg_int])
df5

pltCorSim05 <- df5 %>%
  dplyr::mutate(Sig = case_when(gene %in% svgSigPos ~ "SigPos",
                         gene %in% svgSigNeg ~ "SigNeg",
                         .default = "NotSig")) %>%
  dplyr::filter(Sig %in% c("SigPos", "SigNeg")) %>%
  ggplot2::ggplot() + 
  ggplot2::geom_point(ggplot2::aes(x = correlation, y = similar, color = Sig), alpha = 0.25, size = 1) +
  #xlim(c(0,1)) +
  ggplot2::xlim(NA,1) +
  #ylim(c(0,1)) +
  ggplot2::coord_fixed() +
  #viridis::scale_color_viridis(discrete = TRUE) +
  ggplot2::scale_color_manual(values = c("SigPos" = "green", "SigNeg" = "blue")) +
  ggplot2::theme_classic() + 
  ggplot2::labs(x = "Correlation" , y = "Similarity", 
       #title = "SVGs with minPercentCells = 0.05"
  )


# histCorSim05 <- ggplot(df5, aes(x = similar)) +
#   geom_histogram(binwidth = .10) +
#   theme_classic()

pltCorSim05
# histCorSim05




```















