---
title: "Getting started with STcompare"
author:
- Vivien Jiang
- Kalen Clifton
- Jean Fan
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with STcompare}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.5,
  out.width = "100%"
)
```

# Introduction

Spatial transcriptomics (ST) measures gene expression at known locations in a
tissue. When two tissues are compared, for example a diseased section and a
healthy control, the usual differential expression analysis compares the mean
expression of each gene and ignores where the gene is expressed. A gene can
then look unchanged although its spatial pattern changed, or look changed
although its pattern is the same and only its level differs.

STcompare compares two structurally matched tissues location by location.
For every gene it answers two questions:

| Question | Test | Output |
|---|---|---|
| Is the gene's spatial **pattern** the same in both samples? | Spatial correlation: Pearson's correlation of the two samples over matched pixels, with a permutation p-value that accounts for spatial autocorrelation | `r`, `p`, `padj` |
| Is the gene's expression **level** the same at matched locations? | Spatial similarity: the share of matched pixels where the two values are within a fold change of each other | `similarity` (descriptive, no p-value) |

The two tests are complementary: a gene can have the same pattern at a
different level (high correlation, low similarity), or the same level with a
different pattern.

This tutorial uses small simulated data that ships with the package and runs in
under a minute. The method is described in Clifton et al. (2026),
[doi:10.1093/bioinformatics/btag644](https://doi.org/10.1093/bioinformatics/btag644).

```{r setup, message = FALSE}
library(STcompare)
library(SpatialExperiment)
library(ggplot2)
library(patchwork)
```

# Simulated kidneys

`speKidney` is a list of three simulated single-cell datasets, `A`, `B` and
`C`, each a `SpatialExperiment` with one gene called `Gene`, its counts in the
assay `counts`, and the x-y coordinates of the cells in `spatialCoords()`.

```{r data}
data(speKidney)
speKidney$A
```

```{r cells, fig.height = 3, fig.alt = "Maps of the cells of the simulated kidneys A, B and C, coloured by the expression of Gene: A and C share a pattern and B has the opposite pattern."}
cells <- do.call(rbind, lapply(c("A", "B", "C"), function(s) {
  data.frame(spatialCoords(speKidney[[s]]),
             expression = assay(speKidney[[s]], "counts")["Gene", ],
             sample = s)
}))
ggplot(cells, aes(x, y, colour = expression)) +
  geom_point(size = 0.4) +
  facet_wrap(~sample) +
  coord_equal() +
  scale_colour_viridis_c() +
  theme_void()
```

A and B have the same expression levels but opposite patterns, and A and C have
the same pattern at different levels. Comparing the distributions of the
expression values, which is what a mean-based comparison does, misses both
differences:

```{r boxplot, fig.width = 4, fig.height = 3, out.width = "60%", fig.alt = "Box plots of the expression of Gene in A, B and C: A and B have the same distribution, and C is higher."}
ggplot(cells, aes(sample, expression)) +
  geom_boxplot() +
  labs(x = "Sample", y = "Expression") +
  theme_classic()
```

A and B are indistinguishable by their means, and A and C differ only by a
fold change. STcompare's two tests separate these cases.

# Prepare the input

STcompare compares the same locations in two samples, so the samples must
first be put on a common grid:

1. **Align** the tissues, so that the same coordinates point to the same
   structure in both samples. [STalign](https://jef.works/STalign/) does this
   for sections that are not already aligned; the simulated kidneys are.
2. **Rasterize** both samples **together**, in one call of
   `SEraster::rasterizeGeneExpression()`. This aggregates cells or spots into
   pixels of a shared grid, so that a pixel name means the same location in
   every sample. Samples rasterized in separate calls generally do not share
   their grid, and `compareSpatial()` refuses pixels whose coordinates differ.

```{r rasterize}
rast <- SEraster::rasterizeGeneExpression(speKidney, assay_name = "counts",
                                          resolution = 0.2, fun = "mean",
                                          square = FALSE)
rast$A
```

The result is a list of `SpatialExperiment` objects, one per sample, whose
columns are hexagonal pixels and whose assay `pixelval` holds the mean
expression of the cells in each pixel. Pixels are shared where both samples
have cells.

```{r rasterplot, fig.height = 3, fig.alt = "The three kidneys rasterized onto one grid of hexagonal pixels, coloured by the mean expression of Gene."}
SEraster::plotRaster(rast$A, plotTitle = "A") +
  SEraster::plotRaster(rast$B, plotTitle = "B") +
  SEraster::plotRaster(rast$C, plotTitle = "C")
```

# Compare two samples

`compareSpatial()` runs both tests for every gene in both samples. It takes
two `SpatialExperiment` objects, or a named list of two, whose names label the
samples in the output. Here we compare A with B.

```{r ab}
ab <- compareSpatial(rast[c("A", "B")], nThreads = 2)
ab
```

The printed header summarises the settings and the counts; the table has one
row per gene. `nThreads` sets the number of threads; the results do not
depend on it, and `options(STcompare.nThreads = 8)` sets it for a whole
session. In an interactive session, `compareSpatial()` also shows a progress
line with an estimate of the remaining time while the permutations run.

A and B are strongly negatively correlated (r = `r round(ab$r, 3)`): none of
the `r ab$nPermutations` null correlations was as extreme, so p is the smallest
possible value, 1/`r sprintf("%d", ab$nPermutations + 1L)`. `r round(100 * ab$similarity)`% of
their matched pixels are within two-fold of each other. Now A and C:

```{r ac}
ac <- compareSpatial(rast[c("A", "C")], nThreads = 2)
as.data.frame(ac)[, c("gene", "r", "p", "padj", "nPermutations",
                      "similarity", "dissimilarityX", "dissimilarityY")]
```

A and C are positively correlated (r = `r round(ac$r, 3)`), yet none of their
pixels are within two-fold: in `r round(100 * ac$dissimilarityY)`% of the
pixels kept for the similarity, C is more than twice as high as A.

## Reading the result

| Column | Meaning |
|---|---|
| `gene`, `nPixels` | The gene and the number of shared pixels. |
| `r` | Pearson's correlation of the gene's values in the two samples over the shared pixels. |
| `pNaive` | The p-value of `cor.test()`, which assumes that pixels are independent. Neighbouring pixels are not independent, so this value is far too small; it is shown for reference only (see below). |
| `p` | The permutation p-value of the spatial correlation test. It accounts for the spatial autocorrelation of both samples: the null correlations come from surrogates of one sample that keep its spatial autocorrelation but are unrelated to the other sample. Use this p-value. |
| `padj` | `p` adjusted for multiple testing across the genes of the call (Benjamini-Hochberg by default, `adjustMethod`). |
| `pX`, `pY` | Descriptive p-values of the two directions: surrogates of the first sample (X) correlated with the second, and surrogates of the second (Y) correlated with the first. `p` combines them. |
| `nPermutations`, `stop` | How many permutations the gene used, and why it stopped: `"exceedances"` when it stopped early because its p-value is clearly not small, `"limit"` when it reached the `nPermutations` argument. |
| `deltaStarMedianX`, `deltaStarMedianY` | The median smoothing bandwidth (delta star) chosen for the surrogates of each sample. |
| `deltaGridEdge` | `TRUE` when most surrogates chose the smallest or the largest bandwidth of the grid, a hint to extend the `delta` argument. |
| `similarity` | The share of the kept pixels whose values are within a fold change of each other, \|log2(y / x)\| <= `foldChange` (1, two-fold, by default). |
| `dissimilarityX`, `dissimilarityY` | The share of kept pixels more than `foldChange` higher in the first sample (X) and in the second (Y). The three shares add up to 1. The similarity compares fold changes, so it needs values that are not negative: it is `NA` for genes with negative values. |
| `nPixelsSimilarity`, `thresholdX`, `thresholdY` | The number of pixels kept for the similarity: those above the gene's expression threshold in either sample, `thresholdX` or `thresholdY` (its 5% quantile by default). |
| `status`, `message` | `"ok"`; `"skipped"`, with the reason, for a gene that is not tested for correlation because it is constant or detected in few pixels of a sample (see `minDetected` in `?compareSpatial`); or `"failed"`, with the reason, for a gene that could not be tested (for example because of missing values). |

`as.data.frame()` turns the result into a plain data frame, and `summary()`
gives the counts:

```{r summary}
summary(ac)
```

## Adaptive permutations: genes that are clearly not significant stop early

The permutation p-value is computed from up to `nPermutations` (10,000 by
default) surrogate pairs, but a gene stops as soon as `exceedances` (10 by
default) null correlations are at least as extreme as the observed one. A
gene whose patterns are unrelated stops after a few dozen permutations,
while a significant gene runs to the limit, which sets the smallest possible
p-value, `1 / (nPermutations + 1)`. For example, two independently simulated
patterns from `simRanPatternRasts`:

```{r early}
data(simRanPatternRasts)
indep <- compareSpatial(simRanPatternRasts[[1]], simRanPatternRasts[[2]],
                        nThreads = 2)
as.data.frame(indep)[, c("r", "p", "nPermutations", "stop")]
```

This gene stopped after `r indep$nPermutations` permutations, with p =
`r signif(indep$p, 2)`, while A and B used all `r ab$nPermutations`. The
article [How STcompare works](how-STcompare-works.html) explains why the
p-value stays valid when genes stop at different times.

# Visualize a comparison

The rasterized maps show the patterns. `plotCorrelationGeneExp()` plots the
values of a gene in the two samples at the shared pixels, with its correlation
and adjusted p-value:

```{r correlation-plots, fig.height = 3.5, fig.alt = "Scatter plots of the values of Gene at the shared pixels: A against B falls along a line with negative slope, A against C along a line with positive slope."}
plotCorrelationGeneExp(rast[c("A", "B")], ab, "Gene") +
  plotCorrelationGeneExp(rast[c("A", "C")], ac, "Gene")
```

For the pixel-level view of the similarity, `spatialSimilarity()` computes the
same similarity as `compareSpatial()` and also returns the class of every
pixel, which `linearRegression()` and `pixelClass()` plot:

```{r similarity, fig.height = 3.5, message = FALSE, warning = FALSE, fig.alt = "Left: values of Gene in A against B at each pixel, coloured by whether they are within two-fold. Right: map of the pixels coloured by the same classes."}
simAB <- spatialSimilarity(rast[c("A", "B")])
simAC <- spatialSimilarity(rast[c("A", "C")])
linearRegression(simAB, "Gene") + pixelClass(simAB, "Gene")
```

In the scatter plot (left), each point is a pixel, with its value in A on the
x axis and in B on the y axis. Pixels between the dashed lines are within
two-fold (blue); the others are higher in A (yellow) or in B (red). The map
(right) shows where these pixels are. For A and C, every kept pixel is higher
in C:

```{r similarityAC, fig.height = 3.5, message = FALSE, warning = FALSE, fig.alt = "The same plots for A and C: every kept pixel is more than two-fold higher in C."}
linearRegression(simAC, "Gene") + pixelClass(simAC, "Gene")
```

# Why permutation p-values? A calibration check

Neighbouring pixels have similar values: the patterns are spatially
autocorrelated. Two unrelated but smooth patterns can therefore have a sizeable
correlation by chance, much more often than `cor.test()` expects from
independent observations. `simRanPatternRasts` holds 100 independently
simulated, spatially autocorrelated patterns rasterized onto one grid, so any
two of them are unrelated. We compare 50 disjoint pairs, (1, 2), (3, 4), and
so on, with a different seed for each pair, and collect the results.

```{r calibration}
pairs <- matrix(seq_along(simRanPatternRasts), nrow = 2)
calibration <- do.call(rbind, lapply(seq_len(ncol(pairs)), function(k) {
  res <- compareSpatial(simRanPatternRasts[[pairs[1, k]]],
                        simRanPatternRasts[[pairs[2, k]]],
                        tests = "correlation", nPermutations = 1000,
                        seed = k, nThreads = 2, verbose = FALSE)
  as.data.frame(res)
}))
nrow(calibration)
summary(calibration$nPermutations)
```

Most pairs stopped after a few dozen permutations, so the 50 tests take a few
seconds. The null hypothesis is true for every pair, so a calibrated test
gives p < 0.05 for about 5% of them:

```{r calibration-counts}
c(naive = sum(calibration$pNaive < 0.05),
  permutation = sum(calibration$p < 0.05))
c(naive = sum(p.adjust(calibration$pNaive, "BH") < 0.05),
  permutation = sum(p.adjust(calibration$p, "BH") < 0.05))
```

`r sum(calibration$pNaive < 0.05)` of 50 naive p-values are below 0.05, and
`r sum(p.adjust(calibration$pNaive, "BH") < 0.05)` pairs remain "significant"
after the Benjamini-Hochberg adjustment. The permutation test gives
`r sum(calibration$p < 0.05)` of 50, in line with the 2.5 expected by chance:
the one-sided binomial p-value of seeing at least that many is
`r format.pval(binom.test(sum(calibration$p < 0.05), 50, 0.05, alternative = "greater")$p.value, digits = 2)`,
against
`r format.pval(binom.test(sum(calibration$pNaive < 0.05), 50, 0.05, alternative = "greater")$p.value, digits = 2)`
for the naive test. After the adjustment, the permutation test gives
`r sum(p.adjust(calibration$p, "BH") < 0.05)`.

```{r calibration-plot, fig.height = 3, fig.alt = "Histograms of the p-values of 50 unrelated pairs: the naive p-values pile up near 0, the permutation p-values do not."}
pvalues <- rbind(
  data.frame(test = "naive (cor.test)", p = calibration$pNaive),
  data.frame(test = "permutation (compareSpatial)", p = calibration$p))
ggplot(pvalues, aes(p)) +
  geom_histogram(breaks = seq(0, 1, 0.05), colour = "white") +
  geom_hline(yintercept = 50 * 0.05, linetype = "dashed") +
  facet_wrap(~test) +
  labs(x = "p-value of 50 unrelated pairs", y = "Pairs") +
  theme_classic()
```

Under the null hypothesis, p-values are uniformly distributed (dashed line).
The naive p-values pile up near 0; the permutation p-values do not. Many
permutation p-values are close to 1: a pair whose patterns are clearly
unrelated reaches 10 exceedances within its first few permutations and stops
with p = 10 / L, for example 1 after 10 permutations.

# Migrating from the legacy functions

The functions of earlier versions are still available, unchanged in their
interface. They are computed by the same compiled engine and reproduce the
published analyses exactly. `compareSpatial()` is the recommended entry point
for new analyses.

| Earlier function | `compareSpatial()` |
|---|---|
| `spatialCorrelationGeneExp(list(A = a, B = b))` | `compareSpatial(list(A = a, B = b), tests = "correlation")` |
| `spatialCorrelationGeneExpIterPermutations()` (100, then 1000 permutations for promising genes) | adaptive permutations, the default (`exceedances = 10`, `nPermutations = 10000`) |
| `spatialSimilarity(list(a, b))` | `compareSpatial(a, b, tests = "similarity")` |
| `correlationCoef`, `pValueNaive` | `r`, `pNaive` |
| `pValuePermuteX`, `pValuePermuteY` (each adjusted across genes); significant when both are below 0.05 | `p` = the larger of the two directions' p-values, and `padj`; significant when `padj` < 0.05 |
| `deltaStarMedianX`, `deltaStarMedianY` | the same |
| `percentSimilarity`, `percentDissimilarityX`, `percentDissimilarityY` | `similarity`, `dissimilarityX`, `dissimilarityY` |
| `numPixelInThresh`, `t1`, `t2` | `nPixelsSimilarity`, `thresholdX`, `thresholdY` |

The differences in the results:

- `r`, `pNaive` and the similarity scores are identical.
- The permutation p-values differ within Monte Carlo error: `compareSpatial()`
  draws independent permutations for each gene and direction (the earlier
  functions share one set across genes), stops early for genes that are
  clearly not significant, and adjusts one combined p-value per gene. On the
  published kidney and brain analyses the two approaches agree on the
  significance of 95-98% of the genes (see
  [Parameters, performance and reproducibility](parameters-performance-reproducibility.html)).
- `compareSpatial()` does not test genes detected in only a few of the shared
  pixels of either sample (`minDetected`): the surrogates would give them
  p-values that are too small.
- `compareSpatial()` checks that the shared pixels have the same coordinates,
  uses the genes present in both samples, and reports genes that are skipped or
  cannot be tested in its `status` and `message` columns instead of stopping.

# Next steps

- [How STcompare works](how-STcompare-works.html): the tests step by step.
- [Parameters, performance and reproducibility](parameters-performance-reproducibility.html):
  choosing the resolution, the smoothing grid and the number of permutations;
  threads and run times; seeds.
- Two case studies run the published analyses on real data in a minute or two:
  [acute kidney injury (10x Visium)](https://jef.works/STcompare/articles/acute-kidney-injury-10x-visium-rasterized.html)
  and [MERFISH vs Visium in the mouse brain](https://jef.works/STcompare/articles/brain-MERFISH-10x-visium.html).

# Session information

```{r session}
sessionInfo()
```
