---
title: "How STcompare works"
output:
  rmarkdown::html_vignette:
    math_method: mathml
vignette: >
  %\VignetteIndexEntry{How STcompare works}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.5,
  out.width = "100%"
)
```

This article walks through what `compareSpatial()` computes for one gene,
step by step, with small examples from the data that ship with the package:

1. [Matched pixels](#matched-pixels): the two samples are compared pixel by pixel.
2. [Pearson's correlation](#pearsons-correlation-and-why-the-naive-p-value-fails),
   and why its usual p-value fails for spatial data.
3. [Spatially autocorrelated surrogates](#a-null-distribution-that-keeps-the-spatial-autocorrelation)
   with the gene's own values, which give a valid null distribution.
4. [Sequential permutation p-values](#sequential-p-values-stop-early-when-the-answer-is-clear)
   that stop early when the answer is clear.
5. [Both directions, one p-value per gene](#two-directions-one-p-value) and the
   adjustment for multiple testing.
6. [Rarely detected genes](#rarely-detected-genes): why the surrogates take the
   gene's own values.
7. [Spatial similarity](#spatial-similarity): the expression levels at matched pixels.

```{r setup, message = FALSE}
library(STcompare)
library(SpatialExperiment)
library(ggplot2)
library(patchwork)
data(speKidney)
data(simRanPatternRasts)
```

# Matched pixels

STcompare compares the same locations in two samples. After the tissues are
aligned, both samples are rasterized together onto one grid of pixels, here
hexagons, so that a pixel name refers to the same location in both:

```{r rasterize}
rast <- SEraster::rasterizeGeneExpression(speKidney, assay_name = "counts",
                                          resolution = 0.2, fun = "mean",
                                          square = FALSE)
shared <- intersect(colnames(rast$A), colnames(rast$B))
length(shared)
```

For a gene, the comparison uses two vectors over the shared pixels: its
values in the first sample, `x`, and in the second, `y`.

```{r vectors}
x <- assay(rast$A, "pixelval")["Gene", shared]
y <- assay(rast$B, "pixelval")["Gene", shared]
coords <- spatialCoords(rast$A)[shared, ]
```

To draw maps of values on these pixels, we put them into a copy of the
rasterized object and use `SEraster::plotRaster()`:

```{r mapper, fig.height = 3, fig.alt = "Maps of the values of Gene on the pixels shared by kidneys A and B: the two patterns are opposite."}
mapValues <- function(values, title) {
  spe <- rast$A[, shared]
  assay(spe, "pixelval") <- matrix(values, nrow = 1,
                                   dimnames = list("Gene", shared))
  SEraster::plotRaster(spe, plotTitle = title, showLegend = FALSE)
}
mapValues(x, "x: A") + mapValues(y, "y: B")
```

# Pearson's correlation, and why the naive p-value fails

The statistic is Pearson's correlation of `x` and `y` over the shared pixels:

```{r r}
cor(x, y)
```

Its usual test, `cor.test()`, assumes that the pixels are independent
observations. In a tissue they are not: neighbouring pixels have similar
values, so a smooth pattern carries far less independent information than its
number of pixels suggests, and two unrelated smooth patterns often have a
sizeable correlation. `simRanPatternRasts` holds 100 independently simulated,
spatially autocorrelated patterns on one grid. The correlations of all 4950
pairs of them are spread much more widely than the correlations of
independent pixels would be:

```{r null-r, fig.height = 3, fig.alt = "Histogram of the correlations of 4950 pairs of unrelated simulated patterns, much wider than the red curve of the correlation of independent pixels."}
values <- lapply(simRanPatternRasts, function(s) assay(s, "pixelval")[1, ])
pairs <- utils::combn(length(values), 2)
pairR <- apply(pairs, 2, function(ij) {
  common <- intersect(names(values[[ij[1]]]), names(values[[ij[2]]]))
  c(r = cor(values[[ij[1]]][common], values[[ij[2]]][common]),
    pixels = length(common))
})
r <- pairR["r", ]
pixels <- pairR["pixels", ]
# the p-value of cor.test() for each pair
naiveP <- 2 * pt(-abs(r) * sqrt((pixels - 2) / (1 - r^2)), df = pixels - 2)
n <- median(pixels)
ggplot(data.frame(r = r), aes(r)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, fill = "grey70") +
  stat_function(fun = function(r) dnorm(r, sd = 1 / sqrt(n - 3)),
                colour = "red") +
  labs(x = "Correlation of two unrelated patterns", y = "Density") +
  theme_classic()
```

The red curve is the distribution of `r` for `r n` independent pixels, with a
standard deviation of `r round(1 / sqrt(n - 3), 3)`; the observed standard
deviation is `r round(sd(r), 3)`. The patterns behave as if they had about
`r round(1 / sd(r)^2 + 3)` independent observations rather than `r n`. As a
result, `cor.test()` gives p < 0.05 for `r round(100 * mean(naiveP < 0.05))`%
of these unrelated pairs instead of 5%. STcompare therefore builds the null
distribution of `r` from surrogate patterns with the same spatial
autocorrelation.

# A null distribution that keeps the spatial autocorrelation

STcompare uses the method of Viladomat et al. (2014). To test the correlation
of `x` with `y`, it generates surrogates of `x`: random patterns that have the
same spatial autocorrelation as `x`, and the same values, but no relation to
`y`. Each surrogate is made in five steps.

**1. Shuffle** the values of `x` across the pixels. This keeps the values but
destroys the spatial pattern, and with it the autocorrelation.

```{r shuffle}
set.seed(1)
shuffled <- x[sample.int(length(x))]
```

**2. Smooth** the shuffled values with a Gaussian kernel whose window holds a
share \(\delta\) of the pixels (a nearest-neighbour bandwidth). A small
\(\delta\) gives a rough pattern and a large one a smooth pattern. The smoother
is the local-constant fit of `locfit::locfit(kern = "gauss", nn = delta)`,
evaluated on locfit's adaptive tree, reimplemented in C++.

**3. Match the variogram.** The variogram \(\gamma(h)\) is half the mean
squared difference between the values of two pixels at distance \(h\); it
describes how fast similarity decays with distance. STcompare computes it up
to the 25% quantile of the pixel distances (`maxDistPrctile`; with more than
1000 pixels, on a random subset of 1000 pixels, for speed), fits the
variogram of `x` as a linear function of the variogram of the smoothed values,
\(\gamma_x(h) \approx \beta_0 + \beta_1 \gamma_\delta(h)\), and forms the
surrogate \(\sqrt{|\beta_1|}\, x_\delta + \sqrt{|\beta_0|}\, Z\), where \(Z\) is
Gaussian noise. Rescaling matches the slope of the variogram and the noise its
intercept (the nugget).

**4. Choose \(\delta\).** Steps 2 and 3 are done for every \(\delta\) of the grid
(`delta`, by default 0.01, 0.05 and 0.1 to 0.9), and the \(\delta\) whose
surrogate's variogram is closest to the variogram of `x` (smallest residual sum
of squares), delta star, gives the surrogate.

**5. Remap the values.** The smoothed, rescaled surrogate has values that are
close to normally distributed, whatever the distribution of `x`. Its values are
therefore replaced by the values of `x` themselves, placed in the surrogate's
rank order: the pixel with the smallest surrogate value gets the smallest value
of `x`, and so on (the amplitude adjustment of Theiler et al. 1992;
`surrogate = "remap"`, the default). The surrogate keeps the spatial
arrangement of step 3 and has exactly the distribution of values of `x`, zeros
included. [Rarely detected genes](#rarely-detected-genes) shows why this
matters; `surrogate = "gaussian"` skips this step and gives the surrogates of
the legacy functions.

The legacy function `viladomatCorrelation()` exposes the surrogates of steps
1 to 4. With a single \(\delta\) it returns the surrogate for that bandwidth,
and with the same seed it starts from the same shuffle as above; step 5 is one
line of R:

```{r surrogates, fig.height = 2.6, fig.alt = "Maps of x, of its shuffled values, and of remapped surrogates smoothed with increasing bandwidths: the shuffled values have no pattern, and larger bandwidths give smoother surrogates, all with the values of x."}
xyData <- cbind(x, y, coords)
grid <- c(0.01, 0.05, seq(0.1, 0.9, 0.1))
smoothed <- function(delta) {
  viladomatCorrelation(xyData, delta = delta, maxDistPrctile = 0.25,
                       nPermutations = 1, seed = 1)
}
remap <- function(surrogate, values) {
  remapped <- surrogate
  remapped[order(surrogate)] <- sort(values)
  remapped
}
deltaStar <- smoothed(grid)$deltaStar
deltaStar
shown <- unique(c(0.01, 0.05, deltaStar, 0.6))
gaussian <- lapply(shown, function(d) smoothed(d)$permutations[, 1])
fields <- c(list(x = x, shuffled = shuffled),
            lapply(gaussian, remap, values = x))
names(fields)[-(1:2)] <- ifelse(shown == deltaStar, "delta* ", "delta ")
names(fields)[-(1:2)] <- paste0(names(fields)[-(1:2)], shown)
wrap_plots(Map(mapValues, fields, names(fields)), nrow = 1)
```

Every map has the same values, those of `x`, arranged differently: shuffled,
or smoothed with increasing bandwidths.

The variograms show why delta star is `r deltaStar` for this surrogate. Delta
star is chosen in step 4, from the variograms of the smoothed, rescaled
surrogates; the remapping is a monotone change of the amplitudes, which alters
the variogram only a little, so the variograms of the remapped surrogates tell
the same story. We compute them with a small helper, in equal distance bins up
to the 25% quantile of the distances (STcompare's own computation follows
`geoR::variog()`):

```{r variograms, fig.height = 3.5, fig.alt = "Variograms of x (black points), of the shuffled values (flat) and of the remapped surrogates: the surrogate at delta star follows the variogram of x most closely. A dashed line shows the surrogate at delta star before the remapping."}
variogram <- function(values, coords, maxDist, nBins = 12) {
  d <- as.matrix(dist(coords))
  halfSq <- outer(values, values, function(a, b) (a - b)^2 / 2)
  pair <- upper.tri(d) & d <= maxDist
  bin <- cut(d[pair], seq(0, maxDist, length.out = nBins + 1))
  v <- data.frame(distance = as.numeric(tapply(d[pair], bin, mean)),
                  semivariance = as.numeric(tapply(halfSq[pair], bin, mean)))
  v[!is.na(v$distance), ]
}
maxDist <- quantile(dist(coords), 0.25)
curves <- do.call(rbind, Map(function(v, label) {
  data.frame(variogram(v, coords, maxDist), field = label)
}, fields, names(fields)))
curves$field <- factor(curves$field, levels = names(fields))
target <- curves[curves$field == "x", ]
beforeRemap <- variogram(gaussian[[which(shown == deltaStar)]], coords, maxDist)
ggplot(curves, aes(distance, semivariance, colour = field)) +
  geom_line(linewidth = 0.8) +
  geom_line(data = beforeRemap, colour = "black", linetype = "dashed") +
  geom_point(data = target, colour = "black", size = 2) +
  coord_cartesian(ylim = c(0, 1.5 * max(target$semivariance))) +
  scale_colour_manual(values = c("black", "grey60",
                                 "#E69F00", "#56B4E9", "#009E73", "#CC79A7")) +
  labs(x = "Distance", y = "Semivariance", colour = NULL) +
  theme_classic()
```

The variogram of `x` (black points) rises with distance: nearby pixels are
similar. The shuffled values have a flat variogram. A small bandwidth leaves a
rough pattern whose variogram rises too fast and flattens, a large one an
over-smoothed pattern whose variogram rises too slowly, and delta star follows
the variogram of `x` closely, before the remapping (dashed) and after it.

**6. Null correlations.** The correlation of each surrogate of `x` with `y` is
one draw from the null distribution of `r`. The same is done in the other
direction, with surrogates of `y` correlated with `x`. `keepNulls = TRUE` keeps
these null correlations:

```{r nulls, fig.height = 3, fig.alt = "Histograms of the null correlations of the surrogates of A and of B; the observed correlation, a red line, lies far outside both."}
ab <- compareSpatial(rast[c("A", "B")], nPermutations = 999, nThreads = 2,
                     keepNulls = TRUE)
nulls <- attr(ab, "details")$Gene
ggplot(data.frame(r = c(nulls$nullX, nulls$nullY),
                  direction = rep(c("surrogates of A", "surrogates of B"),
                                  each = length(nulls$nullX))), aes(r)) +
  geom_histogram(bins = 40, fill = "grey70") +
  geom_vline(xintercept = ab$r, colour = "red") +
  facet_wrap(~direction) +
  labs(x = "Null correlation", y = "Permutations") +
  theme_classic()
```

The observed correlation of A and B (red line) lies far outside both null
distributions. The nulls are wide, because unrelated smooth patterns can
correlate strongly, but `r round(ab$r, 2)` is out of their reach.

# Sequential p-values: stop early when the answer is clear

With \(B\) permutations, the permutation p-value of a direction is
\((b + 1) / (B + 1)\), where \(b\) counts the null correlations at least as
extreme as the observed one, \(|r_{null}| \ge |r|\). The smallest possible
value is \(1 / (B + 1)\), so a small p-value needs many permutations; a large
p-value does not. After 30 permutations with 15 exceedances, a gene is clearly
not significant, and 10,000 more permutations would only refine an
uninteresting number.

`compareSpatial()` therefore uses the sequential rule of Besag and Clifford
(1991): a gene draws permutations until `exceedances` (\(h\), 10 by default) null
correlations are at least as extreme as the observed one, or until
`nPermutations` (\(n\), 10,000 by default). Its p-value is

- \(h / L\) if the \(h\)-th exceedance came at permutation \(L\), and
- \((b + 1) / (n + 1)\), the usual value, if it reached \(n\) with \(b < h\)
  exceedances.

Besag and Clifford showed that stopping early costs no validity: if the
observed correlation is distributed like the null correlations, as it is when
the null hypothesis is true and the surrogates reproduce the null distribution,
\(P(p \le \alpha) \le \alpha\) for every \(\alpha\), as for a fixed
number of permutations, so it can be adjusted for multiple testing like any
other p-value. (Where the surrogates do not reproduce the null distribution, as
surrogates without step 5 do not for rarely detected genes, no number of
permutations helps; see [Rarely detected genes](#rarely-detected-genes).) Its
relative precision is
about \(1 / \sqrt{h}\) (about 30% for \(h = 10\)), which is ample for p-values
that are far from significance, while significant genes still get the full
resolution of \(n\) permutations. An unrelated gene costs on average about
\(h (1 + \ln(n / h))\) permutations, around 80 for the defaults.

The figure follows three comparisons permutation by permutation: two
unrelated patterns, a weakly correlated pair made by mixing two of the
simulated patterns, and kidneys A and C. We run 2000 permutations of each
without stopping (`exceedances = Inf`) and count the exceedances as they
accumulate.

```{r sequential, fig.height = 3.2, fig.alt = "Running counts of exceedances against the permutation number for unrelated patterns, weakly related patterns and kidneys A and C, with the stopping line at 10 exceedances."}
f1 <- simRanPatternRasts[[9]]
f2 <- simRanPatternRasts[[10]]
px <- intersect(colnames(f1), colnames(f2))
z1 <- as.numeric(scale(assay(f1, "pixelval")[1, px]))
z2 <- as.numeric(scale(assay(f2, "pixelval")[1, px]))
mixed <- f2[, px]
assay(mixed, "pixelval") <- matrix(0.3 * z1 + sqrt(1 - 0.3^2) * z2, nrow = 1,
                                   dimnames = list(rownames(f2), px))
examples <- list(
  "unrelated patterns" = list(simRanPatternRasts[[1]], simRanPatternRasts[[2]]),
  "weakly related patterns" = list(f1[, px], mixed),
  "kidneys A and C" = rast[c("A", "C")])
fixed <- lapply(examples, function(e) {
  compareSpatial(e[[1]], e[[2]], tests = "correlation", nPermutations = 2000,
                 exceedances = Inf, keepNulls = TRUE, nThreads = 2,
                 verbose = FALSE)
})
running <- do.call(rbind, Map(function(res, label) {
  nl <- attr(res, "details")[[1]]
  rbind(data.frame(b = seq_along(nl$nullX), direction = "x permuted",
                   exceedances = cumsum(abs(nl$nullX) >= abs(res$r)),
                   example = label),
        data.frame(b = seq_along(nl$nullY), direction = "y permuted",
                   exceedances = cumsum(abs(nl$nullY) >= abs(res$r)),
                   example = label))
}, fixed, names(fixed)))
running$example <- factor(running$example, levels = names(examples))
adaptive <- lapply(examples, function(e) {
  compareSpatial(e[[1]], e[[2]], tests = "correlation", nPermutations = 2000,
                 nThreads = 2, verbose = FALSE)
})
stops <- data.frame(example = factor(names(adaptive), levels = names(examples)),
                    L = sapply(adaptive, function(res) res$nPermutations))
ggplot(running, aes(b, exceedances, colour = direction)) +
  geom_step() +
  geom_hline(yintercept = 10, linetype = "dashed") +
  geom_vline(data = stops, aes(xintercept = L), colour = "grey40") +
  scale_x_log10() +
  coord_cartesian(ylim = c(0, 25)) +
  facet_wrap(~example) +
  labs(x = "Permutation", y = "Exceedances so far", colour = NULL) +
  theme_classic() +
  theme(legend.position = "bottom")
```

The dashed line is \(h = 10\), and the grey line marks where the adaptive run
stopped (`nPermutations` of the adaptive result). The unrelated patterns reach
10 exceedances after `r stops$L[1]` permutations and the weakly related pair
after `r stops$L[2]`; kidneys A and C never exceed and run to the limit:

```{r sequential-table}
data.frame(example = names(adaptive),
           r = sapply(adaptive, function(res) round(res$r, 3)),
           nPermutations = sapply(adaptive, function(res) res$nPermutations),
           stop = sapply(adaptive, function(res) res$stop),
           p = sapply(adaptive, function(res) signif(res$p, 3)),
           row.names = NULL)
```

Because each permutation's random numbers depend only on the seed, the gene
and the permutation number, the adaptive run computed exactly the first
\(L\) permutations of the fixed run: it is the same experiment, stopped early.

# Two directions, one p-value

Surrogates of `x` keep the autocorrelation of `x` but not of `y`, and the
reverse for surrogates of `y`. STcompare runs both directions on the same
permutation numbers, in lockstep, and stops a gene as soon as either direction
reaches \(h\) exceedances. The gene's p-value `p` is the larger of the two
directions' sequential p-values:

- if one direction reached \(h\) exceedances at permutation \(L\), `p` is
  \(h / L\) (the other direction, had it continued alone, could only have
  given a smaller value);
- otherwise `p` is \((\max(b_x, b_y) + 1) / (n + 1)\).

The larger of two valid p-values is itself valid, whatever the dependence
between them, so `p` needs no assumption about how the two directions relate.
Taking the larger value formalizes the rule of the original analyses, which
called a gene significant when both directions were. `pX` and `pY`,
\((b + 1) / (L + 1)\) for each direction at the common \(L\), are descriptive:
when a gene stops early, the direction that did not reach \(h\) has an
inflated value.

Across the genes of a call, `padj` adjusts `p` for multiple testing
(Benjamini-Hochberg by default). With \(G\) genes, the \(k\)-th smallest p-value
must be at most \(\alpha k / G\), so the smallest possible p-value,
\(1 / (n + 1)\), matters when few genes are significant among many: see
[Parameters, performance and reproducibility](parameters-performance-reproducibility.html).

# Rarely detected genes

Step 5 exists because of genes detected in a handful of pixels. Without it, the
surrogates are smoothed and mixed with Gaussian noise, so their values are
close to normally distributed (`surrogate = "gaussian"`, the surrogates of the
legacy functions and of the published analyses). That suits a gene expressed
over much of the tissue, but not a gene detected in a handful of pixels: the
correlation of two such genes is decided by the few pixels where both are
detected. Two unrelated genes that happen to share a detected pixel have a
large correlation, and such coincidences happen far more often than the nearly
normal values of gaussian surrogates produce extreme correlations. For
example, two unrelated genes on the pixels of kidneys A and B, each detected in
3 pixels, one of which they share:

```{r sparse}
withValues <- function(values) {
  spe <- rast$A[, shared]
  assay(spe, "pixelval") <- matrix(values, nrow = 1,
                                   dimnames = list("Gene", shared))
  spe
}
sparseX <- sparseY <- numeric(length(shared))
sparseX[c(10, 100, 200)] <- c(3, 8, 5)
sparseY[c(100, 150, 250)] <- c(6, 2, 4)
sparse <- list(withValues(sparseX), withValues(sparseY))
gaussian <- compareSpatial(sparse, tests = "correlation", nPermutations = 999,
                           surrogate = "gaussian", minDetected = 0,
                           nThreads = 2, verbose = FALSE)
as.data.frame(gaussian)[, c("r", "p", "nPermutations")]
```

None of the 999 gaussian surrogate pairs comes close to this correlation, so p
is the smallest possible value. These genes have no spatial pattern, so
shuffling the pixels of one of them gives the exact null distribution, and it
shows how often such a coincidence happens by chance:

```{r sparse-shuffle}
set.seed(2)
shuffled <- replicate(1e5, cor(sample(sparseX), sparseY))
exact <- mean(abs(shuffled) >= abs(gaussian$r) * (1 - 1e-9))
exact
```

(The tolerance counts the shuffles that give exactly the observed correlation,
which `cor()` rounds differently for each arrangement of the values.) A
correlation this large happens by chance about once in `r round(1 / exact)`
shuffles, so the gaussian surrogates' p-value, `r signif(gaussian$p, 2)`, is
`r round(exact / gaussian$p)` times too small; with the default
`nPermutations = 10000` it would be `r round(exact * 10001)` times too small.
Remapped surrogates have the gene's own values. For a gene without spatial
structure the ranks of its surrogates are close to random, so its remapped
surrogates are nearly plain shuffles of its values, and the test is close to
the exact permutation test above, whatever the gene's distribution. The
defaults give a p-value of the right size:

```{r sparse-remap}
remapped <- compareSpatial(sparse, tests = "correlation", nThreads = 2,
                           verbose = FALSE)
c(remapped = remapped$p, exact = exact)
```

```{r sparse-remap-text, include = FALSE}
L <- remapped$nPermutations
remapText <- if (remapped$stop == "exceedances") {
  sprintf("stopped after %d permutations with 10 exceedances, so its p-value is 10 / %d",
          L, L)
} else {
  sprintf("reached the limit of %d permutations with %d exceedances, so its p-value is (%d + 1) / %d",
          L, round(remapped$p * (L + 1)) - 1, round(remapped$p * (L + 1)) - 1, L + 1)
}
```

The remapped gene `r remapText`, where the exact value predicts
`r round(exact * L, 1)` exceedances. The remapped p-value errs on the safe
side: the surrogate of a gene without spatial structure keeps a little
smoothness, because the variogram matching takes the absolute value of a slope
that is pure noise, so its null correlations are spread slightly more widely
than those of plain shuffles. (Many rearrangements of a sparse gene give
exactly the observed correlation; as in the exact test, such ties count as
exceedances, within the rounding of `cor()`.)

With gaussian surrogates, `compareSpatial()` does not test genes detected in
fewer than \(\sqrt{N}\) of the \(N\) shared pixels of either sample (argument
`minDetected`), here `r ceiling(sqrt(length(shared)))` of `r length(shared)`
pixels: the departure from the surrogates grows roughly as \(N / (k_x k_y)\)
for genes detected in \(k_x\) and \(k_y\) pixels, and the filter removes the
genes for which it is largest, although genes detected in more pixels but
still mostly zero can get p-values that are somewhat too small.

```{r sparse-skipped}
skipped <- compareSpatial(sparse, tests = "correlation", surrogate = "gaussian",
                          nThreads = 2, verbose = FALSE)
as.data.frame(skipped)[, c("p", "status", "message")]
```

With remapped surrogates, the default, no such filter is needed. In
simulations on the pixels of the kidney and brain case studies, remapped
surrogates kept \(P(p \le \alpha) \le \alpha\) for independent genes detected
in anywhere from 1% to 100% of the pixels, where gaussian surrogates were
anti-conservative below about 10%, and they had the same power on correlated
fields (`bench/calibration-results.md` in the source repository).
`minDetected` therefore defaults to 0 with remapped surrogates: every gene
that is not constant in a sample is tested, a gene without spatial structure
costs about ten permutations, and the remapping itself, one sort of the
\(N\) values per surrogate, adds about 7% to the run time.

# Spatial similarity

The correlation compares patterns and ignores the levels. The spatial
similarity compares the levels at matched pixels. For each gene:

1. **Threshold.** The threshold of each sample is the gene's `minQuantile`
   quantile (5% by default) over the shared pixels. A pixel is kept when its
   value is above the threshold in either sample: pixels where the gene is
   barely expressed in both samples would otherwise count as similar.
2. **Log ratio.** On the kept pixels, zeros are replaced by \(10^{-4}\) and
   \(l = \log_2(y / x)\).
3. **Classify.** A pixel is similar if \(|l| \le\) `foldChange` (1 by default,
   within two-fold), higher in `x` if \(l <\) `-foldChange`, and higher in `y` if
   \(l >\) `foldChange`.
4. **Score.** `similarity` is the share of similar pixels among the kept
   pixels, and `dissimilarityX` and `dissimilarityY` the shares of the other two
   classes. If fewer than a share `minPixels` (10%) of the pixels are kept, the
   scores are `NA`.

These steps take a few lines of R for kidneys A and B, and give the value
reported by `compareSpatial()`:

```{r similarity-by-hand}
keep <- x > quantile(x, 0.05) | y > quantile(y, 0.05)
x0 <- ifelse(x == 0, 1e-4, x)
y0 <- ifelse(y == 0, 1e-4, y)
l <- log2(y0[keep] / x0[keep])
c(byHand = mean(abs(l) <= 1), compareSpatial = ab$similarity)
```

```{r similarity-plot, fig.height = 3, fig.alt = "Histogram of the log2 ratios of B to A at the kept pixels; the shaded band between -1 and 1 holds the similar pixels."}
ggplot(data.frame(l = l), aes(l)) +
  annotate("rect", xmin = -1, xmax = 1, ymin = -Inf, ymax = Inf,
           fill = "steelblue", alpha = 0.2) +
  geom_histogram(bins = 40, fill = "grey40") +
  labs(x = "log2(B / A) of the kept pixels", y = "Pixels") +
  theme_classic()
```

The shaded band holds the similar pixels. The similarity is a description of
the data and has no p-value. Because it compares levels, it needs values on a
linear scale that are comparable between the samples (for example counts per
million), not log-transformed values.

# Summary

For each gene, `compareSpatial()`:

1. takes the gene's values on the pixels shared by the two samples, and skips
   the correlation test if the gene is constant in a sample (or, with
   gaussian surrogates, detected in too few pixels);
2. computes Pearson's `r` (and, for reference, the naive `pNaive`);
3. draws surrogates of each sample that keep its spatial autocorrelation
   (shuffle, smooth with each \(\delta\), match the variogram, keep delta star)
   and its values (remap), and correlates them with the other sample;
4. stops when either direction has `exceedances` null correlations at least as
   extreme as `r`, or at `nPermutations`, and reports the larger of the two
   directions' sequential p-values as `p`;
5. adjusts `p` across genes (`padj`);
6. computes the similarity of the levels at the kept pixels.

All genes share the expensive parts that depend only on the pixel coordinates
(the smoothing operators and the variogram bins), and the permutations run on
`nThreads` threads.

# References

- Clifton K, Jiang V, dos Santos Peixoto R, Singh S, Matsuura R, Rabb H, Fan J
  (2026). STcompare: comparative spatial transcriptomics data analysis of
  structurally matched tissues to characterize differentially spatially
  patterned genes. *Bioinformatics* 42(9):btag644.
  [doi:10.1093/bioinformatics/btag644](https://doi.org/10.1093/bioinformatics/btag644)
- Viladomat J, Mazumder R, McInturff A, McCauley DJ, Hastie T (2014). Assessing
  the significance of global and local correlations under spatial
  autocorrelation: a nonparametric approach. *Biometrics* 70(2):409-418.
  [doi:10.1111/biom.12139](https://doi.org/10.1111/biom.12139)
- Theiler J, Eubank S, Longtin A, Galdrikian B, Farmer JD (1992). Testing for
  nonlinearity in time series: the method of surrogate data. *Physica D*
  58(1-4):77-94.
  [doi:10.1016/0167-2789(92)90102-S](https://doi.org/10.1016/0167-2789(92)90102-S)
- Besag J, Clifford P (1991). Sequential Monte Carlo p-values. *Biometrika*
  78(2):301-304. [doi:10.1093/biomet/78.2.301](https://doi.org/10.1093/biomet/78.2.301)
- Benjamini Y, Hochberg Y (1995). Controlling the false discovery rate: a
  practical and powerful approach to multiple testing. *Journal of the Royal
  Statistical Society B* 57(1):289-300.
  [doi:10.1111/j.2517-6161.1995.tb02031.x](https://doi.org/10.1111/j.2517-6161.1995.tb02031.x)
