---
title: "Parameters, performance and reproducibility"
output:
  rmarkdown::html_vignette:
    math_method: mathml
vignette: >
  %\VignetteIndexEntry{Parameters, performance and reproducibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article explains how to choose the settings of `compareSpatial()`, what
determines its run time, and what makes its results reproducible. It assumes
the concepts of [How STcompare works](how-STcompare-works.html). The examples
use the simulated kidneys that ship with the package.

```{r setup, message = FALSE}
library(STcompare)
library(SpatialExperiment)
data(speKidney)
rast <- SEraster::rasterizeGeneExpression(speKidney, assay_name = "counts",
                                          resolution = 0.2, fun = "mean",
                                          square = FALSE)
```

# The settings at a glance

| Setting | Default | What it controls | When to change it |
|---|---|---|---|
| Rasterization `resolution` | set in SEraster | The pixel size, hence the number of shared pixels \(N\), the run time and the spatial detail | Larger than the misalignment of the sections, smaller than the structures of interest |
| `assay` | the first assay | The values compared | Use a normalized, linear-scale assay (for example CPM) for the similarity |
| `genes` | the genes in both samples | Which genes are compared | To compare whole transcriptomes, choose the genes with a spatial pattern in both samples |
| `minDetected` | \(\sqrt{N}\) of the \(N\) shared pixels | Which genes are tested for correlation: those detected in at least this share of the pixels of both samples | Raise it (for example to 0.1) when testing many sparsely detected genes; 0 tests every gene that is not constant |
| `delta` | 0.01, 0.05, 0.1, ..., 0.9 | The candidate smoothing bandwidths of the surrogates | When many genes have `deltaGridEdge = TRUE` |
| `nPermutations` | 10000 | The largest number of permutations per gene, hence the smallest p-value, \(1 / (n + 1)\) | Smaller for a quick look when many genes are significant; larger when few genes are significant among many |
| `exceedances` | 10 | When a gene stops early, hence the precision of large p-values | Rarely; `Inf` gives a fixed number of permutations |
| `maxDistPrctile` | 0.25 | The range of distances over which the variograms are matched | Rarely |
| `foldChange` | 1 | The similarity band, \|log2(y / x)\| <= `foldChange` (1 is two-fold) | `log2(1.5)` for 1.5-fold |
| `minQuantile`, `minPixels` | 0.05, 0.1 | Which pixels count for the similarity, and the least share of pixels kept | For genes expressed in a small part of the tissue |
| `adjustMethod` | `"BH"` | The adjustment of `p` across the genes of the call | Any method of `p.adjust()` |
| `seed` | 0 | The random permutations | To check that a conclusion does not depend on the random draws |
| `nThreads` | `getOption("STcompare.nThreads", 1)` | The number of threads | Set it to the number of cores available |

# Preparing the input

**Joint rasterization.** Both samples must be rasterized in one call of
`SEraster::rasterizeGeneExpression()`, after alignment, so that they share
their pixels. `compareSpatial()` uses the pixels whose names are in both
samples and stops if their coordinates differ.

**Resolution.** The resolution sets the number of shared pixels \(N\). Larger
pixels smooth over small misalignments and noise, but merge structures; smaller
pixels resolve more detail, cost more run time per permutation and, when a
pixel holds only one cell or spot, carry the noise of single measurements. As a
guide, the case studies use pixels of about 10 Visium spots (kidney, 311 shared
pixels) and of one Visium spot (brain, 2170 shared pixels).

**Aggregation and normalization.** Either aggregate the counts with
`fun = "sum"` and normalize the pixels, for example to counts per million
(CPM), as in the kidney case study, or normalize the cells or spots first and
aggregate them with `fun = "mean"`, as in the brain case study. The correlation
can also be computed on log-transformed values. The similarity compares fold
changes and needs a linear scale.

**Genes.** A gene that is constant in a sample (for example never detected)
cannot be tested for correlation and gets `status = "skipped"`. So does a gene
detected in fewer than \(\sqrt{N}\) of the \(N\) shared pixels of either
sample (`minDetected`; 18 of 311 pixels, 47 of 2170): its correlation is decided
by the few pixels where both samples detect it, and the surrogates, whose values
are close to normally distributed, would give it a p-value that is far too small
(see [How STcompare works](how-STcompare-works.html#rarely-detected-genes)).
Genes detected in more pixels but mostly zero can still get p-values that are
somewhat too small, so when comparing whole transcriptomes, test the genes that
have a spatial pattern in both samples, as the case studies do, or raise
`minDetected`, for example to 0.1. On the whole AKI raster (32,285 genes on 311
pixels), the default tests 13,500 genes and skips 14,383 that are constant in a
sample and 4,402 that are detected in fewer than 18 pixels.

# The smoothing grid: `delta`

Each surrogate is smoothed with every bandwidth \(\delta\) of the grid, and the
best match of the variogram, delta star, is kept. The default grid adds 0.01
and 0.05 to 0.1, ..., 0.9: with the shorter grid, the smallest value was often
the best one in the published analyses, a sign that the best bandwidth was
smaller still.

`compareSpatial()` drops bandwidths whose window would hold fewer than 2
pixels (\(\lfloor N\delta \rfloor < 2\)), with a message. With 150 pixels, 0.01
is dropped:

```{r dropped}
pixels <- intersect(colnames(rast$A), colnames(rast$B))[1:150]
small <- compareSpatial(rast$A[, pixels], rast$B[, pixels],
                        nPermutations = 199, nThreads = 2)
attr(small, "params")$deltaGrid
```

`deltaGridEdge = TRUE` flags genes for which more than half of the surrogates,
in either direction, chose the smallest or the largest bandwidth, and a
message counts them. Their p-values remain valid, but the surrogates may
reproduce their autocorrelation less well. If many genes are flagged, add
values beyond the edge, for example 0.005 when the smallest value is chosen
(it needs \(N \ge 400\)). Each bandwidth adds to the run time, the small ones on
many pixels the most.

# Permutations: `nPermutations` and `exceedances`

A gene stops when either direction has `exceedances` null correlations at
least as extreme as the observed one, or at `nPermutations`.

**`nPermutations` sets the resolution of small p-values.** A gene that reaches
the limit has \(p = (b + 1) / (n + 1)\), at least \(1 / (n + 1)\). With the
Benjamini-Hochberg adjustment of \(G\) genes, the gene of rank \(k\) is
significant at level \(\alpha\) only if \(p \le \alpha k / G\). If you expect at
least \(k\) significant genes, a limit of

\[ n \ge \frac{G}{\alpha k} - 1 \]

lets genes without any exceedance reach significance:

```{r resolution}
G <- c(100, 1000, 10000)
k <- c(1, 10, 100)
needed <- outer(G, k, function(G, k) pmax(0, ceiling(G / (0.05 * k) - 1)))
dimnames(needed) <- list(genes = G, "significant genes expected" = k)
needed
```

The default, 10,000, suffices for 1000 genes with at least 2 significant
genes, or for 10,000 genes with at least 20. When many genes are significant,
as in both case studies, 1000 gives the same significant genes 6 to 7 times
faster. If genes reach the limit without any exceedance and are still not
significant after the adjustment, `compareSpatial()` says so in a message: a
larger `nPermutations` would resolve them.

**`exceedances` sets the precision of p-values that are not small.** The
relative standard error of a sequential p-value is about
\(1 / \sqrt{h}\) for \(h\) exceedances, about 30% for the default 10. A
gene whose patterns are unrelated costs on average about
\(h (1 + \ln(n / h))\) permutations, around 80 for the defaults, instead of
\(n\). A larger `exceedances` gives more precise large p-values at a
proportional cost; `exceedances = Inf` gives every gene exactly
`nPermutations` permutations and \(p = \max(p_X, p_Y)\).

# The variogram range: `maxDistPrctile`

The variograms that the surrogates must match are computed up to the
`maxDistPrctile` quantile (25% by default) of the distances between the shared
pixels. Pairs of pixels far apart are fewer, their variogram is noisier, and
it reflects the shape of the tissue more than the spatial autocorrelation of
the gene. A smaller value matches only the short-range autocorrelation; a
larger one also the longer-range structure. The default suits most data;
`compareSpatial()` stops with an error that suggests a larger value when too
few pixel pairs fall in the range to compute a variogram.

# Similarity: `foldChange`, `minQuantile` and `minPixels`

`foldChange` is on the log2 scale: 1 means within two-fold, `log2(1.5)` within
1.5-fold. Each gene gets a threshold in each sample, its `minQuantile`
quantile over the shared pixels (or `t1` and `t2` in `spatialSimilarity()`), and
a pixel counts when it is above the threshold in either sample. If fewer than
a share `minPixels` of the pixels count, the similarity is `NA`. For a gene
expressed in a small region, lower `minPixels`; to ignore background levels, raise
`minQuantile`.

# Threads and run time

The run time is proportional to the total number of permutations (the sum of
the `nPermutations` column) times the cost of one permutation in both
directions, divided by the number of threads. All genes share the parts that
depend only on the pixel coordinates, and the permutations of all genes are
spread over `nThreads` threads. The cost of a permutation grows with the
number of pixels and of bandwidths: one permutation in one direction took
about 0.16 ms of one thread for the kidney data (311 pixels, 11 bandwidths) and
1.1 ms for the brain data (2170 pixels, of which the variograms use a random
subset of 1000).

Measured run times of the case studies (Apple M1 Ultra):

| Data | Genes | Shared pixels | `nPermutations` | Threads | Run time |
|---|---:|---:|---:|---:|---:|
| Kidney, control vs AKI | 1046 | 311 | 1000 | 8 | 29 s |
| Kidney, control vs AKI | 1046 | 311 | 1000 | 16 | 15 s |
| Kidney, control vs AKI | 1046 | 311 | 10000 | 16 | 106 s |
| Brain, MERFISH vs Visium | 325 | 2170 | 1000 | 8 | 40 s |
| Brain, MERFISH vs Visium | 325 | 2170 | 1000 | 16 | 21 s |
| Brain, MERFISH vs Visium | 325 | 2170 | 10000 | 16 | 124 s |

The genes that reach the limit dominate: with the default limit, they used
79-86% of all permutations. To predict the run time of a large analysis, run a
random subset of genes with the settings you plan to use and scale the run time
(`attr(res, "runtime")`) by the number of genes.

**Threads do not change the results.** Each permutation's random numbers
depend only on the seed, the gene and the permutation number, so the results
are identical for any number of threads:

```{r threads}
one <- compareSpatial(rast[c("A", "C")], nPermutations = 999, nThreads = 1,
                      verbose = FALSE)
two <- compareSpatial(rast[c("A", "C")], nPermutations = 999, nThreads = 2,
                      verbose = FALSE)
identical(as.data.frame(one), as.data.frame(two))
```

Set the default for a session with `options(STcompare.nThreads = 8)`. The
threads are threads of the compiled code: no R processes are started, so they
work the same on every operating system, and they never call R. Use at most
the number of cores available, and fewer on a shared computer.

**Memory.** `keepNulls = TRUE` keeps two numbers per permutation and
direction (the null correlation and the delta star), up to 320 MB for 1000
genes that all run to 10,000 permutations. Use it for a few genes.

**Progress.** In an interactive session, a progress line shows the share of
the work done, the genes finished, the permutations so far, and an estimate of
the remaining time:

```
compareSpatial:  61% | genes done 640/1046 | 1.9M permutations | elapsed 0:12 | ETA 0:08
```

In a console narrower than the line, the elapsed time is left out. At the end,
the line shows the total number of permutations, the sum of the
`nPermutations` column. `progress = FALSE` turns it off and `verbose = FALSE`
the other messages.

# Reproducibility

**Results depend only on the data, the settings and `seed`.** Each gene and
direction has its own random stream, determined by the seed, the gene name,
the direction and the permutation number. A gene therefore gets the same
permutations whatever the number of threads, the order of the genes, or the
other genes in the call; only `padj`, which adjusts across the genes of the
call, depends on the other genes. The permutations are drawn over the shared
pixels in the column order of the first sample, as in the legacy functions:
reordering its pixels, or swapping the two samples, gives other permutations.
A different seed does too, and the p-values then differ within Monte Carlo
error:

```{r seeds}
unrelated <- function(seed) {
  compareSpatial(simRanPatternRasts[[1]], simRanPatternRasts[[2]], seed = seed,
                 tests = "correlation", nThreads = 2, verbose = FALSE)
}
data(simRanPatternRasts)
rbind(seed0 = as.data.frame(unrelated(0))[, c("p", "nPermutations")],
      seed1 = as.data.frame(unrelated(1))[, c("p", "nPermutations")])
```

**The session's random number generator is left unchanged.**
`compareSpatial()` draws its permutations and noise in compiled code. With more
than 1000 shared pixels, it also draws the pixels of the variogram subsample
from `seed` with R's default generator, in a local state that it restores
afterwards. It does not affect the random numbers of the rest of an analysis:

```{r rng}
set.seed(42)
before <- runif(1)
set.seed(42)
invisible(compareSpatial(rast[c("A", "B")], nPermutations = 99,
                         nThreads = 2, verbose = FALSE))
identical(before, runif(1))
```

**Platforms.** The results are deterministic on a given platform. Between
platforms (for example macOS and Linux), the Gaussian noise of the surrogates
can differ in the last bits, and on regular grids of pixels, pixel pairs that
lie exactly at the largest variogram distance can be binned differently. The
surrogates then differ slightly, and p-values can differ within Monte Carlo
error.

**Record the settings.** The result keeps every setting, including the
bandwidths actually used, the seed and the assays, in
`attr(res, "params")`, along with the call and the run time.

# The legacy functions

The functions of earlier versions remain available with their interface and
semantics. They are computed by the same compiled engine as `compareSpatial()`,
much faster than before, and reproduce the published analyses: the same delta
stars, the same null correlations up to rounding, and the same p-values, except
that p-values are now \((b + 1) / (B + 1)\), never 0, and adjusted across genes
as documented.

| Function | What it does | Use it to |
|---|---|---|
| `compareSpatial()` | Both tests for every gene; adaptive permutations; one p-value per gene | Analyse new data |
| `spatialCorrelationGeneExp()` | The correlation test for every gene with a fixed number of permutations; p-values of both directions | Reproduce published results |
| `spatialCorrelationGeneExpIterPermutations()` | The same, with a second round of more permutations for genes that look significant | Reproduce published results |
| `spatialCorrelationGeneExpWithinSample()` | The correlation test for pairs of genes within one sample | Compare genes with each other |
| `spatialCorrelation()`, `viladomatCorrelation()` | The correlation test for vectors of values and coordinates | Use the test outside of `SpatialExperiment` objects |
| `spatialSimilarity()` | The similarity, with the class of every pixel | Plot pixel classes with `linearRegression()` and `pixelClass()` |

Re-running the six published analyses with the legacy functions on macOS
(arm64), where they were computed, gives identical delta stars and null
correlations within \(10^{-13}\), and takes seconds instead of hours (16
threads of an Apple M1 Ultra, against the authors' runs on 20 to 22 workers):

| Analysis | Genes | Shared pixels | Run time | Authors' run time |
|---|---:|---:|---:|---:|
| Kidney, control vs AKI (100 permutations) | 1046 | 311 | 1.9 s | not recorded |
| Kidney, control vs AKI (100, then 1000 permutations) | 1046 | 311 | 14 s | not recorded |
| MERFISH replicates, affine alignment | 483 | 1299 | 40 s | 7.65 h |
| MERFISH replicates, STalign alignment | 483 | 1371 | 40 s | 16.8 h |
| Brain, MERFISH vs Visium | 325 | 2170 | 13 s | 1.78 h |
| Brain cell types | 16 | 2174 | 0.9 s | 6.95 min |

The legacy functions share one sequence of permutations across genes and
directions (the published results depend on it), use a fixed number of
permutations per round, and report a p-value per direction. `compareSpatial()`
uses independent permutations for each gene and direction, stops early for
genes that are clearly not significant, skips rarely detected genes, and
reports one p-value per gene. On the published analyses, the two agree on the
significance of 98% of the kidney genes and 95% of the brain genes, counting a
gene as significant with the legacy function when both of its adjusted p-values
are below 0.05:

| Analysis | Significant with both | Legacy function only | `compareSpatial()` only | Skipped by `compareSpatial()` |
|---|---:|---:|---:|---:|
| Kidney, 1046 genes | 722 | 8 (1 skipped) | 16 | 7 |
| Brain, 325 genes | 122 | 6 (4 skipped) | 10 | 28 |

The tested genes classified differently are near the significance threshold
(adjusted p-values between 0.015 and 0.07), where Monte Carlo error and the
different adjustments (one p-value per gene rather than one per direction, over
the tested genes) decide. The five skipped genes that the legacy function found
significant are detected in fewer than \(\sqrt{N}\) pixels of a sample: Kcnt1
in 17 of the 311 pixels of the AKI section, and four brain genes with weak
correlations (\(r\) = 0.06 to 0.10).
