---
title: "Acute Kidney Injury (10x Visium) Application"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{acute-kidney-injury-10x-visium}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  dpi = 72,
  dev = "png",
  fig.path = "getting-started-with-STcompare-figures/"

)
```

```{r setup}
library(STcompare)
library(rhdf5)
library(Matrix)
library(ggplot2)
library(dplyr)
library(patchwork)
library(BiocGenerics)

```

# Introduction 

In the [`Getting Started with STcompare`](https://jef.works/STcompare/articles/getting-started-with-STcompare.html) tutorial, we demonstrated that `STcompare` works on both simulated and real ST data. We next sought to apply it to compare healthy and diseased datasets to discover potentially disease-relevant changes in spatial gene expression patterns. To this end, we focus on a common and serious disease, acute kidney injury (AKI), by comparing two previously published murine kidney sections assayed by spot-resolution 10x Visium: a section from a kidney 24 hours post initial AKI injury and a section from a normal kidney as a control. 

This tutorial corresponds with Figures 2f-j from the [`STcompare paper`](https://doi.org/10.1101/2025.11.21.689847). 

The files are stored as a H5 file on Zenodo. We are assuming here that the counts and the position matrices are well formed and correspond properly with one another. 

# Load H5 files

```{r load-data}

# The Visium mouse kidney datasets can be found Zenodo: https://zenodo.org/records/17676992
# IL3 - ischemia acute kidney injury dataset 
# NL3 - control dataset 

# AKI counts ###################################################################
zenodo_url <- "https://zenodo.org/records/17676992/files/IL3_filtered_feature_bc_matrix.h5?download=1"
temp_h5 <- tempfile(fileext = ".h5")
download.file(zenodo_url, temp_h5, mode = "wb")

# this is a list of the internal groups and datasets stored in the HDF5 file 
# we are interested in the "matrix/barcodes" dataset 
rhdf5::h5ls(temp_h5)

# Read the CSC-encoded matrix components from the HDF5 file
file_barcodes <- as.character(rhdf5::h5read(temp_h5, "matrix/barcodes"))
barcodes_matrix <- rhdf5::h5read(temp_h5, "matrix") # stores the sparse matrix components 

# convert CSC encoding into an R dgCMatrix
aki_counts <- Matrix::sparseMatrix(
  dims = barcodes_matrix$shape,
  i = as.numeric(barcodes_matrix$indices),
  p = as.numeric(barcodes_matrix$indptr),
  x = as.numeric(barcodes_matrix$data),
  index1 = FALSE
)

colnames(aki_counts) <- file_barcodes
rownames(aki_counts) <- as.data.frame(barcodes_matrix[["features"]])$name

unlink(temp_h5)
head(aki_counts)
```

```{r load-data2}
# Control counts ###############################################################

zenodo_url <- "https://zenodo.org/records/17676992/files/NL3_filtered_feature_bc_matrix.h5?download=1"
temp_h5 <- tempfile(fileext = ".h5")
download.file(zenodo_url, temp_h5, mode = "wb")

# Read the CSC-encoded matrix components from the HDF5 file
file_barcodes <- as.character(rhdf5::h5read(temp_h5, "matrix/barcodes"))
barcodes_matrix <- rhdf5::h5read(temp_h5, "matrix") # stores the sparse matrix components 

# convert CSC encoding into an R dgCMatrix
control_counts <- Matrix::sparseMatrix(
  dims = barcodes_matrix$shape,
  i = as.numeric(barcodes_matrix$indices),
  p = as.numeric(barcodes_matrix$indptr),
  x = as.numeric(barcodes_matrix$data),
  index1 = FALSE
)

colnames(control_counts) <- file_barcodes
rownames(control_counts) <- as.data.frame(barcodes_matrix[["features"]])$name

unlink(temp_h5)
head(control_counts)
```

```{r load-data3}
# positions ####################################################################

zenodo_url <- "https://zenodo.org/records/17676992/files/IL3_tissue_positions.csv?download=1"
aki_pos <- read.csv(zenodo_url, header = TRUE, stringsAsFactors = FALSE)

zenodo_url <- "https://zenodo.org/records/17676992/files/NL3_tissue_positions.csv?download=1"
ctrl_pos <- read.csv(zenodo_url, header = TRUE, stringsAsFactors = FALSE)

# given the counts the the position matrix 
head(colnames(aki_counts))
head(aki_pos$barcode)
```

```{r load-data4}
# Standardize spot identifiers by removing sample-specific prefixes
# The regex "^[^_]*_" matches from the start (^) through the first underscore:
#   [^_]* = any characters except underscore
#   _     = the first underscore itself
# Original format: "SAMPLE_X_Y" 
# Standardized format: "X_Y"
aki_pos$barcode <- sub("^[^_]*_", "", aki_pos$barcode)
ctrl_pos$barcode <- sub("^[^_]*_", "", ctrl_pos$barcode)

# limit the counts to the spots with positions and make sure they are in the same order

# AKI 
# aki_pos$barcode with tag removed and colnames(aki_counts) are not identical 
identical(aki_pos$barcode, colnames(aki_counts))

# There are spots in aki_counts that are not in aki_pos
setdiff(aki_pos$barcode, colnames(aki_counts))
setdiff(colnames(aki_counts), aki_pos$barcode)

# removing the spots that are in aki_counts but not in aki_pos 
# remove the tag from the position barcodes 
aki_counts <- aki_counts[, colnames(aki_counts) %in% aki_pos$barcode]

# they are now identical, if this was not the case, more alignments 
# step would've been needed 
identical(sub("^[^_]+_", "", aki_pos$barcode), colnames(aki_counts))

# Control 
# The control is also not identical 
identical(ctrl_pos$barcode, colnames(control_counts))

# same case as before, remove the spots that are in the counts, but no in the pos
setdiff(ctrl_pos$barcode, colnames(control_counts))
setdiff(colnames(control_counts), ctrl_pos$barcode)
control_counts <- control_counts[, colnames(control_counts) %in% ctrl_pos$barcode]

# they are now identical
identical(ctrl_pos$barcode, colnames(control_counts))
```

```{r load-data5}
# plot positions ###############################################################

# set the barcode column as the rownames
rownames(ctrl_pos) <- ctrl_pos$barcode
ctrl_pos <- ctrl_pos[,-1]
colnames(ctrl_pos) <- c("x", "y")

rownames(aki_pos) <- aki_pos$barcode
aki_pos <- aki_pos[,-1]
colnames(aki_pos) <- c("x", "y")

ctrl_pos$group <- "Control"
aki_pos$group  <- "AKI"

df <- rbind(ctrl_pos, aki_pos)
pos_plot <- ggplot(df, aes(x = x, y = y, color = group)) +
  geom_point() +
  scale_color_manual(values = c("Control" = "blue", "AKI" = "red")) +
  coord_fixed() +
  labs(x = "x", y = "y", color = "Group") +
  theme_classic()
pos_plot

# rotate the tissue by 90-degrees for consistency with the paper 

ctrl_pos_rot <- transform(ctrl_pos, x = y, y = -x + max(ctrl_pos$x))
aki_pos_rot <- transform(aki_pos, x = y, y = -x + max(aki_pos$x))

df_rot <- rbind(ctrl_pos_rot, aki_pos_rot)
pos_rot_plot <- ggplot(df_rot, aes(x = x, y = y, color = group)) +
  geom_point(size=0.5) +
  scale_color_manual(values = c("Control" = "blue", "AKI" = "red")) +
  coord_fixed() +
  labs(x = "x", y = "y", color = "Group") +
  theme_classic()
pos_rot_plot

```

# Create SpatialExperiment 

Now that it has been verified that both the counts and the position matrices are well formed, the input to `STcompare` is a [`SpatialExperiment`]("https://www.bioconductor.org/packages/release/bioc/html/SpatialExperiment.html"). They are objects that store spatial transcriptomics data, in this case, combining gene expression matrices with the spatial coordinates of each spatial location. 

``` {r create_spatial_experiment1}

# Turn the position and gene expression counts matrix into a spatial experiment 
# with the following assays 
#   counts: raw gene expression counts
#   libnorm: CPM (counts per million) normalization to adjust for sequencing depth

# remove the group column from position 
ctrl_pos_rot <- ctrl_pos_rot[, -3]
aki_pos_rot <- aki_pos_rot[, -3]

AKI_ctrl_SE <- SpatialExperiment::SpatialExperiment(
  assays = list(counts = control_counts,
                libnorm = Matrix::t(Matrix::t(control_counts)/Matrix::colSums(control_counts)*1e6)),
  spatialCoords = as.matrix(ctrl_pos_rot),
  
)

AKI_irl_SE <- SpatialExperiment::SpatialExperiment(
  assays = list(counts = aki_counts,
                libnorm = Matrix::t(Matrix::t(aki_counts)/Matrix::colSums(aki_counts))*1e6),
  spatialCoords = as.matrix(aki_pos_rot)

)

```

Removes rarely-expressed genes that are noisy by filtering to genes that are detected in at least 1% of spots in both samples.

``` {r, create_spatial_experiment2}

# there are 32285 genes in both samples 
dim(SummarizedExperiment::assays(AKI_ctrl_SE)$libnorm) 
dim(SummarizedExperiment::assays(AKI_irl_SE)$libnorm) 

# which rows in the matrix that the gene is expressed in greater than 1% of all spots 
# this has to be true in both datasets to be considered a "good gene" 
good.genes <- names(which(
  (Matrix::rowSums(SummarizedExperiment::assays(AKI_ctrl_SE)$libnorm > 0) / dim(AKI_ctrl_SE)[2]*100 > 1) &
  (Matrix::rowSums(SummarizedExperiment::assays(AKI_irl_SE)$libnorm > 0) / dim(AKI_irl_SE)[2]*100 > 1)))
length(good.genes)

# Reduced the number of genes from 32285 --> 12697 genes 
AKI_ctrl_SE <- AKI_ctrl_SE[good.genes,]
AKI_irl_SE <- AKI_irl_SE[good.genes,]
```

# Identify SVGs 

Of the "good genes" that have expression in at least 1% of spots, we now want to identify which of those genes are then spatially variable (SVG). We use Moran's I to identify SVGs. 

``` {r moransI1}

moransI <- function(SE_temp){
  
  # use MERINGUE to get the neighbor-relationships 
  w <- MERINGUE::getSpatialNeighbors(SpatialExperiment::spatialCoords(SE_temp), filterDist = 1.75)
  par(mfrow=c(1,1))
  MERINGUE::plotNetwork(SpatialExperiment::spatialCoords(SE_temp), w)
  
  # Identify significantly spatially auto-correlated genes
  I <- MERINGUE::getSpatialPatterns(SummarizedExperiment::assays(SE_temp)$libnorm, w)
  
  results.filter <- MERINGUE::filterSpatialPatterns(mat = SummarizedExperiment::assays(SE_temp)$libnorm,
                                          I = I,
                                          w = w,
                                          adjustPv = TRUE,
                                          alpha = 0.05,
                                          minPercentCells = 0.01,
                                          verbose = TRUE, details = TRUE)

  return(results.filter)
}

moransI_ctrl <- moransI(AKI_ctrl_SE)
head(moransI_ctrl)
```

``` {r moransI2}
moransI_irl <- moransI(AKI_irl_SE)
head(moransI_irl)

# "moransI_ctrl" and "moransI_irl" are dataframes these columns: 
# observed: Observed Moran's I statistic measuring spatial autocorrelation
#           Positive values = spatially clustered expression pattern
#           Values range from -1 to 1

# expected: Expected Moran's I under null hypothesis of random distribution
#           Typically close to zero

# sd: Standard deviation of Moran's I under the null hypothesis
#     Used to calculate the test statistic

# p.value: Statistical significance of the spatial pattern
#          Lower values = stronger evidence of non-random spatial expression

# p.adj: Adjusted p-value (FDR-corrected) for multiple testing correction

# minPercentCells: Minimum percent of cells that must be driving spatial pattern as calculated by LISA
```

``` {r moransI3}
# Running Moran's I already filtered the dataframe to only the genes that are statistically significantly autocorrelated 
svg_ctrl <- moransI_ctrl %>% rownames()
svg_irl <- moransI_irl %>% rownames()
svg_int <- intersect(svg_ctrl, svg_irl) # genes that are SVGs in both conditions 
svg_uni <- union(svg_ctrl, svg_irl) # genes that are SVG in one condition but not the other 

# Out of 32285 gene: 
# there are 3389 genes that are SVGs in the control dataset 
length(svg_ctrl)

# there are 4447 genes that are SVGs in the aki dataset  
length(svg_irl)

# there are 2278 genes that are SVGs in both datasets   
length(svg_int)

# there are 5558 genes that are SVGs in either datasets 
length(svg_uni)
```

Unlike previous methods, `STcompare` is to be able to identify SVGs that change in spatial pattern across datasets. Existing SVG analysis methods can identify genes that change from spatially variable to not or vice versa across conditions. However, genes that are classified as spatially variable in both conditions can still vary in the specific spatial pattern in each condition. 

With the SVGs, use STcompare to calculate Pearson's correlation coefficient for genes that were SVGs in both conditions to investigate whether there was a change in the spatial patterning of expression. 

``` {r spatial-correlation1}
# choosing the gene that are SVGs in both conditions 
genes_chosen <- svg_int

# The input is the spatial experiment for the genes that are SVGs in both conditions 
input <- list('AKI_ctrl'= AKI_ctrl_SE[genes_chosen, ], 
             'AKI_irl'= AKI_irl_SE[genes_chosen, ])

# running spatialCorrelationGeneExp 
set.seed(0)
start_time <- Sys.time()
deltaList <- replicate(length(rownames(input[[1]])), c(0.01, 0.05, seq(0.1, 0.9, .1)), simplify = FALSE)
kidneyCorrelation <- STcompare::spatialCorrelationGeneExp(
  input, 
  nPermutations = 1000, # increased the number of permutations to account for stochasticity 
  deltaX = deltaList, 
  deltaY = deltaList, 
  returnPermutations = FALSE, 
  assayName = "libnorm", 
  nThreads = 5
)
end_time <- Sys.time()
print(end_time - start_time) 
```

``` {r spatial-correlation2}
# 527 genes are SVGs that are significantly positively correlated 
kidneyCorrelation %>%
  dplyr::filter(rownames(kidneyCorrelation) %in% svg_int) %>%
  dplyr::filter(pValuePermuteX < 0.05 & pValuePermuteY < 0.05) %>%
  dplyr::filter(correlationCoef > 0) %>% 
  dim()

# saving the significantly positively correlated svg genes 
svgSigPos <- kidneyCorrelation %>%
  dplyr::filter(rownames(kidneyCorrelation) %in% svg_int) %>%
  dplyr::filter(pValuePermuteX < 0.05 & pValuePermuteY < 0.05) %>%  
  dplyr::filter(correlationCoef > 0) %>% 
  rownames()


# 25 genes are SVGs that are significantly negatively correlated 
kidneyCorrelation %>%
  dplyr::filter(rownames(kidneyCorrelation) %in% svg_int) %>%
  dplyr::filter(pValuePermuteX < 0.05 & pValuePermuteY < 0.05) %>%  
  dplyr::filter(correlationCoef < 0) %>% 
  dim()

# saving the significantly negatively correlated svg genes 
svgSigNeg <- kidneyCorrelation %>%
  dplyr::filter(rownames(kidneyCorrelation) %in% svg_int) %>%
  dplyr::filter(pValuePermuteX < 0.05 & pValuePermuteY < 0.05) %>%  
  dplyr::filter(correlationCoef < 0) %>% 
  rownames()

# visualize the significantly positively correlate and significantly negatively correlated svg genes  
# Figure 2J - Rate of Significance for SVGs 
fig_2j <- kidneyCorrelation %>%
  dplyr::mutate(Sig = dplyr::case_when(rownames(kidneyCorrelation) %in% svgSigPos ~ "SigPos",
                         rownames(kidneyCorrelation) %in% svgSigNeg ~ "SigNeg",
                         .default = "")) %>%
  dplyr::mutate(pValueEmpirical = dplyr::case_when(pValuePermuteY > pValuePermuteX ~ pValuePermuteY,
                                     .default = pValuePermuteX)) %>%
  dplyr::mutate(pValueEmpiricalRound = dplyr::case_when(pValueEmpirical == 0 ~ 0.01,
                                    .default = pValueEmpirical)) %>%
  ggplot2::ggplot() + 
  ggplot2::geom_point(ggplot2::aes(x= correlationCoef, y = -log10(pValueEmpiricalRound), color = Sig), alpha = 0.25, size = 1) +
  ggplot2::xlim(NA,1) +
  ggplot2::scale_color_manual(values = c("SigPos" = "green", "SigNeg" = "blue")) +
  ggplot2::theme_classic() +
  ggplot2::geom_hline(yintercept = -log10(0.05), linetype = 'dashed', color = "black")  + 
  ggplot2::labs(x = "Correlation" , y = "-log10(p-value)", title = "Rate of Significance for SVGs")
fig_2j


# Adding a column for the empirical p-value (pValueEmpirical) 
# defined as the larger of the two x and y permuted p-values (greater of pValuePermuteX and pValuePermuteY)
kidneyCorrelationEmpirical <- kidneyCorrelation %>%
  dplyr::mutate(pValueEmpirical = dplyr::case_when(pValuePermuteY > pValuePermuteX ~ pValuePermuteY,
                                     .default = pValuePermuteX))


```

# Spatial Correlation and Fold Change Similarity 

Here we show two examples of the significantly correlated SVG genes. Cbr1 is negatively correlated while Acsm2 is positively correlated.

In Crb1, for the control, gene expression closer to the center of the tissue (inner medulla), while in the AKI sample, gene expression was closer the edge of the issue (cortex). 
1. The **correlation metric** was used to show that there is a significantly negative correlation between the two samples. 
2. The **fold change similarity metric** was used to show that samples are not similar based on fold change. 

In Acsm2, gene expression was concentrated around the edge of the issue (cortex) in both cases, but 
1. The **correlation metric** was used to show that there is a significantly positive correlation between the two samples. 
2. Even though expression pattern remained the same, **fold change similarity metric** was used to show that there is higher overall expression in the control. 


``` {r correlation-plots}

# only the spots that are in both datasets
sharedPixels <- intersect(rownames(SpatialExperiment::spatialCoords(input[[1]])),
                          rownames(SpatialExperiment::spatialCoords(input[[2]])))


# Figure 2g - Visualization of example genes (Cbr1 and Acsm2)

plot_gene_spatial <- function(gene, AKI_ctrl_SE, AKI_irl_SE, sharedPixels) {

  # Control dataframe
  ctrl_df <- data.frame(
    SpatialExperiment::spatialCoords(AKI_ctrl_SE)[sharedPixels, ],
    color = SummarizedExperiment::assay(input[[1]], "libnorm")[gene, sharedPixels]
  )

  # AKI dataframe
  aki_df <- data.frame(
    SpatialExperiment::spatialCoords(AKI_irl_SE)[sharedPixels, ],
    color = SummarizedExperiment::assay(input[[2]], "libnorm")[gene, sharedPixels]
  )

  # Control plot
  ctrl_plot <- ggplot2::ggplot(ctrl_df, ggplot2::aes(x = x, y = y, color = color)) +
    ggplot2::geom_point(size = 0.7, alpha = 1, shape = 18) +
    viridis::scale_color_viridis() +
    ggplot2::coord_fixed() +
    ggplot2::theme_void() +
    ggplot2::ggtitle(paste0("Control: ", gene))

  # AKI plot
  aki_plot <- ggplot2::ggplot(aki_df, ggplot2::aes(x = x, y = y, color = color)) +
    ggplot2::geom_point(size = 0.7, alpha = 1, shape = 18) +
    viridis::scale_color_viridis() +
    ggplot2::coord_fixed() +
    ggplot2::theme_void() +
    ggplot2::ggtitle(paste0("AKI: ", gene))

  return(list(control = ctrl_plot, aki = aki_plot))
}

cbr1_plt <- plot_gene_spatial("Cbr1", AKI_ctrl_SE, AKI_irl_SE, sharedPixels)
acsm2_plt <- plot_gene_spatial("Acsm2", AKI_ctrl_SE, AKI_irl_SE, sharedPixels)

# Figure 2h - example of correlation visualization 
# use the plotCorrelationGeneExp function to generate the correlation plots 

Cbr1_corr <- plotCorrelationGeneExp(
  speList = input, 
  spatialCorrelation = kidneyCorrelation, 
  geneName = "Cbr1", 
  assayName = "libnorm"
  )

Acsm2_corr <- plotCorrelationGeneExp(
  speList = input, 
  spatialCorrelation = kidneyCorrelation, 
  geneName = "Acsm2", 
  assayName = "libnorm"
  )


# Finding the spatial similarity
# If t1 and t2 are null, then the default threshold is the 0.05 quantile of gene expression. 
start_time <- Sys.time()
ss <- STcompare::spatialSimilarity(input, foldChange = 1, assayName = "libnorm", t1 = NULL, t2 = NULL)
end_time <- Sys.time()
print(end_time-start_time)


# Figure 2I - fold change linear regression and spot classification 
acsm2_lr <- STcompare::linearRegression(ss, gene="Acsm2", assayName = "libnorm") + ggplot2::theme_classic() + ggplot2::coord_fixed()
cbr1_lr <- STcompare::linearRegression(ss, gene="Cbr1", assayName = "libnorm") + ggplot2::theme_classic() + ggplot2::coord_fixed()

acsm2_pc <- pixelClass(ss, gene="Acsm2", assayName = "libnorm")
cbr1_pc <- pixelClass(ss, gene="Cbr1", assayName = "libnorm")

cbr1_plt$control + cbr1_plt$aki + Cbr1_corr + cbr1_lr + cbr1_pc
acsm2_plt$control + acsm2_plt$aki + Acsm2_corr + acsm2_lr + acsm2_pc


```


Finally, we compare the correlation and similarity metrics for each gene and showed that the compare the correlation and similarity metrics offer orthogonal insights. 

``` {r fig2k}

# 2k - Comparing Correlation Metric to Similiarty metric 

# create a df of correlation and similarity values for each gene 
sim_corr <- data.frame(gene = ss$similarityTable$gene[ss$similarityTable$gene %in% svg_int],
                  correlation = kidneyCorrelation[svg_int, "correlationCoef" ], 
                  similar = ss$similarityTable$percentSimilarity[ss$similarityTable$gene %in% svg_int])
sim_corr

sim_corr_plt <- sim_corr %>%
  dplyr::mutate(Sig = case_when(gene %in% svgSigPos ~ "SigPos",
                         gene %in% svgSigNeg ~ "SigNeg",
                         .default = "NotSig")) %>%
  dplyr::filter(Sig %in% c("SigPos", "SigNeg")) %>%
  ggplot2::ggplot() + 
  ggplot2::geom_point(ggplot2::aes(x = correlation, y = similar, color = Sig), alpha = 0.25, size = 1) +
  ggplot2::xlim(NA,1) +
  ggplot2::coord_fixed() +
  ggplot2::scale_color_manual(values = c("SigPos" = "green", "SigNeg" = "blue")) +
  ggplot2::theme_classic() + 
  ggplot2::labs(x = "Correlation" , y = "Similarity", 
                title = "Similarity for significantly positively and negatively correlated SVGs")

sim_corr_plt

```















