---
title: "MERFISH Biological Replicates Comparision"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{biological-replicates-example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(STcompare)
library(SpatialExperiment)
library(SEraster)
library(viridis)
library(MERINGUE)
library(patchwork)
library(ggplot2)
library(dplyr)
library(ggrepel)
library(BiocGenerics)
library(BiocParallel)

```

# Introduction 

In the [`Getting Started with STcompare`](https://jef.works/STcompare/articles/getting-started-with-STcompare.html) tutorial, we demonstrated that `STcompare` works on both simulated and real ST data. We next sought to apply it to two biological replicates of coronal slices of the adult mouse brain from different animals but the same location with respect to bregma assayed by the single-cell resolution ST technology MERFISH. Because these are biological replicates, we hypothesized that genes should be significantly positively correlated with high similarity, particularly for spatially variable genes (SVGs) which exhibit coordinated expression variation within tissue structures that likely have correspondence across replicates.

The MERFISH replicates datasets can be found on Zenodo stored as csv.gz files: https://zenodo.org/records/10724029

Dataset S2R3 was aligned to S2R2 in the STalign tutorial: https://jef.works/STalign/notebooks/merfish-merfish-alignment.html 


```{r load-data}

s2r2_url <- "https://zenodo.org/records/10724029/files/STalign_S2R2.csv.gz?download=1"
s2r3_url <- "https://zenodo.org/records/10724029/files/STalign_S2R3_to_S2R2.csv.gz?download=1"

# read the data from the url 
read_csv_gz_url <- function(x, ...) {
  u  <- url(x, open = "rb")
  gz <- gzcon(u, text = TRUE)
  on.exit({
    try(close(gz), silent = TRUE)
    try(close(u),  silent = TRUE)
  }, add = TRUE)

  read.csv(gz, ...)
}

target <- read_csv_gz_url(s2r2_url)
source <- read_csv_gz_url(s2r3_url)

# TODO: what are the column names 
head(target)
head(source)


```

```{r create-spe}

# convert target to spatial experiments

# 1) target ####################################################################
dim(target)

# TODO: 
target[1:5, 1:10]

# extract the cell positions (x, y)
pos_target <- target[, c('x', 'y')]

# TODO: 
rownames(pos_target) <- target$X

# TODO:
# after the third column, the remaining columns are the genes
# making this the counts matrix for spatial expression 
# gene_target <- target[, 4:dim(target)[2]]
gene_target <- target[, 5:dim(target)[2]]
rownames(gene_target) <- target$X

# for spatial experiments, convert from a dataframe to a dgCMatrix
class(gene_target)
target_sparse <- as(t(gene_target), "dgCMatrix")

# format into SpatialExperiment

# TODO: cpm normalize? 
spe_target <- SpatialExperiment::SpatialExperiment(
  assays = list(counts = target_sparse),
  spatialCoords = as.matrix(pos_target)
)

# spe_target <- SpatialExperiment::SpatialExperiment(
#   assays = list(counts = Matrix::t(Matrix::t(target_sparse)/Matrix::colSums(target_sparse)*1e6)),
#   spatialCoords = as.matrix(pos_target)
# )

# 2) source  ###################################################################

dim(source)

# TODO: 
source[1:5, 1:10]

# TODO:
pos_source <- source[, c('STalign_x', 'STalign_y')]
rownames(pos_source) <- source$X
colnames(pos_source) <- c("x", "y")
# gene_source <- source[, 8:dim(source)[2]]

# TODO: 
gene_source <- source[, 9:dim(source)[2]]

rownames(gene_source) <- source$X

# need to convert from data.frame to dgCMatrix
class(gene_source)
source_sparse <- as(t(gene_source), "dgCMatrix")

# TODO: cpm normalize? 
spe_source <- SpatialExperiment::SpatialExperiment(
  assays = list(counts = as(t(gene_source), "dgCMatrix")),
  spatialCoords = as.matrix(pos_source)
)

# spe_source <- SpatialExperiment::SpatialExperiment(
#   assays = list(counts =  Matrix::t(Matrix::t(source_sparse)/Matrix::colSums(source_sparse)*1e6)),
#   spatialCoords = as.matrix(pos_source)
# )



```


``` {r rast}

# Rasterize the Spatial Experiments onto the same coordinate plane 

# make sure the target and the source have the same gene names
# if they are not the same, subset the spatial experiements so that they are the same 
identical(rownames(spe_target), rownames(spe_source))

# turn the spatial experiments into a list so they can be rasterized 
# onto the same coordinate plane 
spe_list <- list(target = spe_target, source = spe_source)

# TODO: is there a reason why we choose 200? 
# TODO: why are we using the mean 
# TODO: multicore? 
# output <- SEraster::rasterizeGeneExpression(
#   spe_list, 
#   resolution = 200, 
#   fun = "mean", 
#   BPPARAM = BiocParallel::MulticoreParam())

output <- SEraster::rasterizeGeneExpression(
  spe_list, 
  resolution = 200, 
  BPPARAM = BiocParallel::SerialParam()
)


# TODO: add title 
plt1 <- SEraster::plotRaster(output[[1]], name = "total counts") + ggplot2::theme_void()
plt2 <- SEraster::plotRaster(output[[2]], name = "total counts") + ggplot2::theme_void()

plt1 + plt2


```

``` {r moranI}


moransI <- function(SE_temp, filterDist = resolution){
  
  # Get neighbor-relationships
  w <- MERINGUE::getSpatialNeighbors(spatialCoords(SE_temp), filterDist = filterDist)
  par(mfrow=c(1,1))
  plotNetwork(spatialCoords(SE_temp), w)
  
  # Identify significantly spatially auto-correlated genes
  start_time <- Sys.time()
  I <- MERINGUE::getSpatialPatterns(as.matrix(assay(SE_temp)), w)
  end_time <- Sys.time()
  print(end_time - start_time)
  
  results.filter <- MERINGUE::filterSpatialPatterns(mat = assay(SE_temp),
                                          I = I,
                                          w = w,
                                          adjustPv = TRUE,
                                          alpha = 0.05,
                                          minPercentCells = 0.01,
                                          verbose = TRUE, details = TRUE)
  
  return(results.filter)
}

start_time <- Sys.time()
moransI_merfish_target <- moransI(output$target, filterDist = resolution)
end_time <- Sys.time()
print(end_time - start_time) #Time difference of 46.42798 secs

start_time <- Sys.time()
moransI_merfish_source <- moransI(output$source, filterDist = resolution)
end_time <- Sys.time()
print(end_time - start_time) #Time difference of 2.000802 hours

svg_merfish_target <- moransI_merfish_target %>% filter(minPercentCells > 0.01) %>% rownames()
svg_merfish_source <- moransI_merfish_source %>% filter(minPercentCells > 0.01) %>% rownames()

svg_int <- intersect(svg_merfish_target, svg_merfish_source)
svg_uni <- union(svg_merfish_target, svg_merfish_source)
non_svg <- rownames(output$target)[!(rownames(output$target) %in% svg_int)]

length(svg_int)
length(non_svg)



```

```{r correlation}

set.seed(0)
start_time <- Sys.time()
deltaList <- replicate(length(rownames(output$target)), c(0.01, 0.05, seq(0.1, 0.9, .1)))
merfishCorrelation <- spatialCorrelationGeneExp(
  output, 
  deltaX = deltaList, 
  deltaY = deltaList, 
  nThreads = 5, 
  nPermutations = 1000 # TODO? 
  )
end_time <- Sys.time()
print(end_time - start_time) #Time difference of 31.29929 mins


# TODO: plot correlation for the two genes 

# TODO: rate of significance plot 

```

```{r similarity}


```


``` {r }

# TODO: similarity vs correlation plot 

```







