---
title: "Acute kidney injury (10x Visium)"
author:
- Kalen Clifton
- Vivien Jiang
- Jean Fan
output: rmarkdown::html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  out.width = "100%"
)
```

This case study compares two mouse kidney sections assayed with 10x Visium: a
kidney 24 hours after ischemic acute kidney injury (AKI) and a kidney after a
sham surgery (control). It looks for genes whose spatial pattern of expression
changed with the injury, and reproduces Figures 2f-k of the
[STcompare paper](https://doi.org/10.1093/bioinformatics/btag644) with
`compareSpatial()`.

The analysis runs live. It downloads 30 MB once (the files are cached for
later runs), and takes a minute or two on a computer with 8 cores, of which the
comparison of 1046 genes takes about half.

```{r setup, message = FALSE}
library(STcompare)
library(SpatialExperiment)
library(ggplot2)
library(patchwork)

# use up to 8 threads, leaving one core free
threads <- min(8L, max(1L, parallel::detectCores() - 1L, na.rm = TRUE))
```

Reading the 10x HDF5 files needs the Bioconductor package rhdf5
(`BiocManager::install("rhdf5")`).

# Download the data

The counts, the spot positions and the positions of the AKI section aligned to
the control section are on Zenodo
([doi:10.5281/zenodo.17676991](https://doi.org/10.5281/zenodo.17676991)). The
helper below downloads a file once into a cache directory and checks its MD5
checksum, so a broken or partial download is never used.

```{r download}
cacheDir <- file.path(tools::R_user_dir("STcompare", which = "cache"),
                      "downloads")

downloadCached <- function(url, file, md5) {
  dir.create(cacheDir, recursive = TRUE, showWarnings = FALSE)
  path <- file.path(cacheDir, file)
  if (file.exists(path) && unname(tools::md5sum(path)) == md5) return(path)
  old <- options(timeout = max(1800, getOption("timeout")))
  on.exit(options(old))
  partial <- paste0(path, ".part")
  utils::download.file(url, partial, mode = "wb", quiet = TRUE)
  if (unname(tools::md5sum(partial)) != md5) {
    unlink(partial)
    stop("The download of ", file, " is incomplete or has changed ",
         "(MD5 mismatch); please try again.", call. = FALSE)
  }
  file.rename(partial, path)
  path
}

zenodo <- "https://zenodo.org/records/19074288/files/"
files <- list(
  aki_counts = downloadCached(
    paste0(zenodo, "IL3_filtered_feature_bc_matrix.h5?download=1"),
    "IL3_filtered_feature_bc_matrix.h5", "a3ea3a2cb4c3e01403d0116ab580c157"),
  control_counts = downloadCached(
    paste0(zenodo, "NL3_filtered_feature_bc_matrix.h5?download=1"),
    "NL3_filtered_feature_bc_matrix.h5", "6793a0f7ceeb082bce383748cfc2805a"),
  control_positions = downloadCached(
    paste0(zenodo, "NL3_tissue_positions.csv?download=1"),
    "NL3_tissue_positions.csv", "24b23088b7a22a53086f9b62b21fed29"),
  aki_aligned = downloadCached(
    paste0("https://zenodo.org/records/19486091/files/",
           "aki_region_onehot_STalign_to_ctrl_region_onehot_affine_only.csv.gz",
           "?download=1"),
    "aki_region_onehot_STalign_to_ctrl_region_onehot_affine_only.csv.gz",
    "adede04898833d385eb52fa2032f2434"))
```

The cache is `tools::R_user_dir("STcompare", "cache")`; set the environment
variable `R_USER_CACHE_DIR` to move it, or delete the directory to free the
space.

# Counts and positions

The counts are 10x HDF5 files, with genes in rows and spots in columns. Gene
symbols are not unique in the 10x feature files (a few symbols belong to two
genes), so `make.unique()` adds a suffix to the second one: STcompare needs
unique gene names.

```{r counts}
read10xH5 <- function(path) {
  h5 <- rhdf5::h5read(path, "matrix")
  counts <- Matrix::sparseMatrix(i = as.integer(h5$indices),
                                 p = as.integer(h5$indptr),
                                 x = as.numeric(h5$data),
                                 dims = as.integer(h5$shape), index1 = FALSE)
  dimnames(counts) <- list(make.unique(as.character(h5$features$name)),
                           as.character(h5$barcodes))
  counts
}
akiCounts <- read10xH5(files$aki_counts)
controlCounts <- read10xH5(files$control_counts)
dim(akiCounts)
dim(controlCounts)
```

The position files list the spots under the tissue, with their row and column
on the Visium array. For the AKI section, the file also has its positions
after alignment to the control section with
[STalign](https://jef.works/STalign/): an affine alignment of the two sections'
cortex, outer medulla and inner medulla, computed in Python.

```{r positions}
controlPositions <- read.csv(files$control_positions, row.names = 1)
akiAligned <- read.csv(files$aki_aligned, row.names = 1)
head(controlPositions, 3)
head(akiAligned[, c("x", "y", "aligned_x", "aligned_y")], 3)

# keep the spots under the tissue
controlCounts <- controlCounts[, rownames(controlPositions)]
akiCounts <- akiCounts[, rownames(akiAligned)]
```

We rotate the sections by 90 degrees so that the kidneys are upright, as in
the paper, and compare the positions before and after the alignment:

```{r alignment, fig.height = 3.5, fig.alt = "Spot positions of the control (blue) and AKI (red) sections before and after the alignment of the AKI section to the control section."}
rotate <- function(x, y) cbind(x = y, y = max(x) - x)
controlXY <- rotate(controlPositions$array_row, controlPositions$array_col)
akiXY <- rotate(akiAligned$x, akiAligned$y)
akiAlignedXY <- rotate(akiAligned$aligned_x, akiAligned$aligned_y)
rownames(controlXY) <- rownames(controlPositions)
rownames(akiXY) <- rownames(akiAlignedXY) <- rownames(akiAligned)

overlay <- function(aki, title) {
  ggplot(rbind(data.frame(controlXY, sample = "control"),
               data.frame(aki, sample = "AKI")), aes(x, y, colour = sample)) +
    geom_point(size = 0.2) +
    scale_colour_manual(values = c(control = "blue", AKI = "red")) +
    coord_equal() +
    labs(title = title, colour = NULL) +
    theme_void() +
    theme(plot.title = element_text(size = 11, hjust = 0.5))
}
overlay(akiXY, "Before alignment") + overlay(akiAlignedXY, "After alignment") +
  plot_layout(guides = "collect")
```

# Rasterize both sections together

A spot of one section has no counterpart at exactly the same position in the
other, so both sections are aggregated onto one grid of hexagonal pixels with
`SEraster::rasterizeGeneExpression()`, called **once** for both, so that a
pixel name means the same location in both. The resolution should be larger
than the remaining misalignment of the two sections and small enough to keep
the structures of interest, such as the thin layers of the medulla. Here a
pixel is 5 array units across and holds about 10 spots; `fun = "sum"` adds the
counts of the spots in a pixel.

```{r rasterize}
control <- SpatialExperiment(assays = list(counts = controlCounts),
                             spatialCoords = controlXY)
aki <- SpatialExperiment(assays = list(counts = akiCounts),
                         spatialCoords = akiAlignedXY)
rast <- SEraster::rasterizeGeneExpression(list(control = control, AKI = aki),
                                          assay_name = "counts",
                                          resolution = 5, fun = "sum",
                                          square = FALSE)
rast$control
summary(rast$control$num_cell)
```

The pixels differ in their total counts, so we normalize each pixel to counts
per million (CPM). The similarity test compares levels, so it needs a linear
scale such as CPM.

```{r cpm, fig.height = 3.5, fig.alt = "The control and AKI sections rasterized onto shared hexagonal pixels, coloured by the total counts of each pixel."}
cpm <- function(m) Matrix::t(Matrix::t(m) / Matrix::colSums(m)) * 1e6
for (s in names(rast)) {
  assay(rast[[s]], "CPM") <- cpm(assay(rast[[s]], "pixelval"))
}
SEraster::plotRaster(rast$control, assay_name = "pixelval",
                     plotTitle = "control", name = "total counts") +
  SEraster::plotRaster(rast$AKI, assay_name = "pixelval",
                       plotTitle = "AKI", name = "total counts")
```

# Choose the genes

STcompare can test every gene, but genes without a spatial pattern in either
section have little to compare. The published analysis tested the genes that
are spatially variable in **both** sections, to find genes whose pattern
changed even though they are patterned in both. It identified them with
Moran's I, as implemented in the GitHub package
[MERINGUE](https://github.com/JEFworks-Lab/MERINGUE), on the CPM of the
pixels:

```{r meringue, eval = FALSE}
# Not run: needs remotes::install_github("JEFworks-Lab/MERINGUE")
moransI <- function(spe, filterDist = 10) {
  w <- MERINGUE::getSpatialNeighbors(spatialCoords(spe), filterDist = filterDist)
  MERINGUE::getSpatialPatterns(assay(spe, "CPM"), w)
}
controlI <- moransI(rast$control)
akiI <- moransI(rast$AKI)
svgGenes <- intersect(rownames(controlI)[controlI$p.adj == 0],
                      rownames(akiI)[akiI$p.adj == 0])
```

To keep this tutorial free of GitHub-only dependencies, STcompare ships the
resulting list: the 1046 genes of the published analysis, spatially variable in
both sections (an adjusted p-value of 0 in both; recomputing them with the
current MERINGUE gives the same list up to 4 genes at the edge of that
criterion).

```{r svg}
svgGenes <- scan(system.file("extdata", "vignette-aki-svg-genes.txt",
                             package = "STcompare"),
                 what = "", comment.char = "#", quiet = TRUE)
length(svgGenes)
head(svgGenes)
```

Without such a selection, most genes of the transcriptome are detected in few
pixels or not at all. `compareSpatial()` does not test a gene that is constant
in a section, or detected in fewer than \(\sqrt{N}\) of the \(N\) shared
pixels of either section (`minDetected`): the correlation of such a gene is
decided by a few pixels, and its p-value would be too small (see
[How STcompare works](https://jef.works/STcompare/articles/how-STcompare-works.html#rarely-detected-genes)).
Here that means 18 of the 311 shared pixels. Counting the shared pixels in
which each gene is detected in the section where it is detected least shows
how many genes a comparison of the whole transcriptome would test:

```{r detected}
shared <- intersect(colnames(rast$control), colnames(rast$AKI))
detected <- sapply(rast, function(r) Matrix::rowSums(assay(r, "pixelval")[, shared] > 0))
table(cut(pmin(detected[, "control"], detected[, "AKI"]), c(-1, 0, 17, Inf),
          labels = c("not detected", "1 to 17 pixels", "18 pixels or more")))
```

Sparse genes above the threshold can still get p-values that are somewhat too
small, which is another reason to test the spatially variable genes.

# Run the comparison

`compareSpatial()` runs the spatial correlation test and the spatial
similarity for every gene, on the pixels shared by the two sections. The named
list `rast` labels the samples: `control` is the first sample (x) and `AKI` the
second (y).

```{r compare}
res <- compareSpatial(rast, assay = "CPM", genes = svgGenes,
                      nPermutations = 1000, nThreads = threads)
res
```

`r sum(res$status == "skipped")` of the spatially variable genes are detected
in fewer than 18 shared pixels of a section and are not tested (status
`"skipped"`).

**Why `nPermutations = 1000`?** The permutations of a gene stop early when its
p-value is clearly not small (here `r sum(res$stop == "exceedances")` of
`r sum(res$status == "ok")` tested genes, after a median of
`r median(res$nPermutations[res$stop == "exceedances"])` permutations), so the
run time is dominated by the significant genes, which run to `nPermutations`.
With 1000, the smallest possible p-value is 1/1001 and this comparison takes
`r round(attr(res, "runtime")[["elapsed"]])` seconds on `r threads` threads.
That resolution is enough here, because many genes are significant: with the
Benjamini-Hochberg adjustment, the \(k\)-th smallest of \(G\) p-values needs to
be at most \(0.05\,k/G\), which is far above 1/1001 for the hundreds of genes
that reach it. The default, `nPermutations = 10000`, is safer when few genes
are significant among many, and takes about 7 times longer for these genes; on
these data it finds exactly the same significant genes. See
[Parameters, performance and reproducibility](https://jef.works/STcompare/articles/parameters-performance-reproducibility.html).

# Interpret the results

We classify the tested genes by the sign of their correlation and their
adjusted p-value:

```{r classes}
tab <- as.data.frame(res)
tab <- tab[tab$status == "ok", ]
tab$class <- factor(ifelse(tab$padj < 0.05,
                           ifelse(tab$r > 0, "positive", "negative"),
                           "not significant"),
                    levels = c("positive", "negative", "not significant"))
table(tab$class)
```

Of the `r nrow(tab)` tested genes that are spatially variable in both sections,
`r sum(tab$class == "positive")` have significantly correlated patterns in the
control and the AKI section, and `r sum(tab$class == "negative")` significantly
anti-correlated patterns. (The published analysis, with the legacy two-stage
permutations, found 707 and 24.) Most of these genes keep their pattern after
the injury; a small group has the opposite pattern.

```{r violin, fig.height = 3.5, fig.alt = "Violin plots of the correlation between the control and AKI sections for significantly positive, significantly negative and not significant genes."}
ggplot(tab, aes(class, r, colour = class)) +
  geom_violin() +
  geom_jitter(width = 0.2, alpha = 0.2, size = 0.6) +
  scale_colour_manual(values = c(positive = "forestgreen", negative = "blue",
                                 "not significant" = "grey50")) +
  labs(x = NULL, y = "Correlation (control vs AKI)") +
  theme_classic() +
  theme(legend.position = "none")
```

Correlation describes the pattern; the similarity describes the levels. Among
the significantly correlated genes, positively and negatively correlated genes
cover the whole range of similarity: the two tests answer different questions.

```{r correlation-vs-similarity, fig.height = 3.5, fig.alt = "Similarity against correlation for the significantly correlated genes: positively and negatively correlated genes cover the whole range of similarity."}
ggplot(subset(tab, class != "not significant"),
       aes(r, similarity, colour = class)) +
  geom_point(alpha = 0.4, size = 1) +
  scale_colour_manual(values = c(positive = "forestgreen", negative = "blue")) +
  labs(x = "Correlation", y = "Similarity", colour = NULL) +
  theme_classic()
```

# Example genes

Two examples illustrate the combinations:

1. a significantly positively correlated gene with the lowest similarity: the
   same pattern at a different level;
2. a significantly negatively correlated gene whose pixels are higher in the
   control about as often as in the AKI section: a different pattern at a
   similar overall level.

```{r examples}
positive <- tab[tab$class == "positive", ]
gene1 <- positive$gene[order(positive$similarity, -positive$r)][1]
negative <- tab[tab$class == "negative", ]
gene2 <- negative$gene[order(abs(negative$dissimilarityX -
                                 negative$dissimilarityY))][1]
tab[c(gene1, gene2), c("gene", "r", "padj", "similarity",
                       "dissimilarityX", "dissimilarityY")]
```

`dissimilarityX` is the share of pixels more than two-fold higher in the
control (x), and `dissimilarityY` in the AKI section (y). We plot the CPM of
the two genes on the shared pixels, and the classes of the pixels from
`spatialSimilarity()`, which `linearRegression()` and `pixelClass()` draw:

```{r example-plots, fig.height = 6.5, message = FALSE, warning = FALSE, fig.alt = "For each of the two example genes: maps of its CPM in the control and AKI sections, a scatter plot of the two at each pixel coloured by similarity class, and a map of the classes."}
shared <- intersect(colnames(rast$control), colnames(rast$AKI))
sim <- spatialSimilarity(list(control = rast$control[c(gene1, gene2), ],
                              AKI = rast$AKI[c(gene1, gene2), ]),
                         assayName = "CPM")
genePanels <- function(gene) {
  (SEraster::plotRaster(rast$control[, shared], assay_name = "CPM",
                        feature_name = gene, plotTitle = paste(gene, "control"),
                        name = "CPM") |
     SEraster::plotRaster(rast$AKI[, shared], assay_name = "CPM",
                          feature_name = gene, plotTitle = paste(gene, "AKI"),
                          name = "CPM")) /
    (linearRegression(sim, gene, assayName = "CPM") |
       pixelClass(sim, gene, assayName = "CPM"))
}
genePanels(gene1)
genePanels(gene2)
```

```{r example-text, include = FALSE}
higher1 <- if (tab[gene1, "dissimilarityY"] > tab[gene1, "dissimilarityX"]) {
  list(share = tab[gene1, "dissimilarityY"], where = "after the injury")
} else {
  list(share = tab[gene1, "dissimilarityX"], where = "in the control")
}
```

`r gene1` has the same pattern in both sections
(r = `r round(tab[gene1, "r"], 2)`), but in
`r round(100 * higher1$share)`% of the kept pixels it is more than two-fold
higher `r higher1$where`. `r gene2` is anti-correlated
(r = `r round(tab[gene2, "r"], 2)`) although
`r round(100 * tab[gene2, "similarity"])`% of its pixels are within two-fold:
its overall level did not change, but where it is highest did. The pixels more
than two-fold higher in the control
(`r round(100 * tab[gene2, "dissimilarityX"])`%, yellow) and after the injury
(`r round(100 * tab[gene2, "dissimilarityY"])`%, red) lie in different regions.

# Session information

```{r session}
sessionInfo()
```
