---
title: "Comparison of MERFISH and Visium for mouse brain"
author:
- Kalen Clifton
- Vivien Jiang
- Jean Fan
output: rmarkdown::html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  out.width = "100%"
)
```

This case study compares two sections of the adult mouse brain assayed with two
different technologies: MERFISH, which measures a targeted panel of genes in
single cells, and 10x Visium, which measures the whole transcriptome in spots of
55 µm. The sections come from different animals and were aligned with
[STalign](https://jef.works/STalign/). We ask how well the spatial patterns of
gene expression, and of cell types, correspond across the two technologies.

The analysis runs live. It downloads 75 MB once (the files are cached for
later runs), and takes about two minutes on a computer with 8 cores.

```{r setup, message = FALSE}
library(STcompare)
library(SpatialExperiment)
library(ggplot2)
library(patchwork)

# use up to 8 threads, leaving one core free
threads <- min(8L, max(1L, parallel::detectCores() - 1L, na.rm = TRUE))
```

Reading the Visium scale factors needs the package jsonlite.

# Download the data

The MERFISH data, slice 2 replicate 3 aligned to the Visium section, are on
Zenodo ([doi:10.5281/zenodo.10724029](https://doi.org/10.5281/zenodo.10724029)),
the Visium data on the
[10x Genomics website](https://www.10xgenomics.com/datasets/adult-mouse-brain-ffpe-1-standard-1-3-0),
and the cell-type annotations of both on Zenodo
([doi:10.5281/zenodo.19582556](https://doi.org/10.5281/zenodo.19582556)). The
helper below downloads a file once into a cache directory and checks its MD5
checksum, so a broken or partial download is never used.

```{r download}
cacheDir <- file.path(tools::R_user_dir("STcompare", which = "cache"),
                      "downloads")

downloadCached <- function(url, file, md5) {
  dir.create(cacheDir, recursive = TRUE, showWarnings = FALSE)
  path <- file.path(cacheDir, file)
  if (file.exists(path) && unname(tools::md5sum(path)) == md5) return(path)
  old <- options(timeout = max(1800, getOption("timeout")))
  on.exit(options(old))
  partial <- paste0(path, ".part")
  utils::download.file(url, partial, mode = "wb", quiet = TRUE)
  if (unname(tools::md5sum(partial)) != md5) {
    unlink(partial)
    stop("The download of ", file, " is incomplete or has changed ",
         "(MD5 mismatch); please try again.", call. = FALSE)
  }
  file.rename(partial, path)
  path
}

tenx <- paste0("https://cf.10xgenomics.com/samples/spatial-exp/1.3.0/",
               "Visium_FFPE_Mouse_Brain/")
celltypes <- "https://zenodo.org/records/19582556/files/"
files <- list(
  merfish = downloadCached(
    "https://zenodo.org/records/10724029/files/STalign_S2R3_to_Visium.csv.gz?download=1",
    "STalign_S2R3_to_Visium.csv.gz", "c092af84ca82a50539f4fdc0f51e8fdc"),
  visium_counts = downloadCached(
    paste0(tenx, "Visium_FFPE_Mouse_Brain_filtered_feature_bc_matrix.tar.gz"),
    "Visium_FFPE_Mouse_Brain_filtered_feature_bc_matrix.tar.gz",
    "017fc078b4aa21e1df39027d0bf3ccad"),
  visium_spatial = downloadCached(
    paste0(tenx, "Visium_FFPE_Mouse_Brain_spatial.tar.gz"),
    "Visium_FFPE_Mouse_Brain_spatial.tar.gz", "a78bb08009c4defd791354dd5c292bfa"),
  type_similarity = downloadCached(
    paste0(celltypes, "STalign_cell_type_transcriptional_correlations.csv.gz?download=1"),
    "STalign_cell_type_transcriptional_correlations.csv.gz",
    "f2a4dc1997ed8ba2344094d534ff64fd"),
  merfish_types = downloadCached(
    paste0(celltypes, "STalign_S2R3_cell_type_annotations.csv.gz?download=1"),
    "STalign_S2R3_cell_type_annotations.csv.gz", "085e49b9bb8f0693d5854d2013b351dc"),
  visium_types = downloadCached(
    paste0(celltypes, "STalign_Visium_cell_type_annotations.csv.gz?download=1"),
    "STalign_Visium_cell_type_annotations.csv.gz", "897985b732e45a7b4dbee3a64cd357e8"))
```

The cache is `tools::R_user_dir("STcompare", "cache")`; set the environment
variable `R_USER_CACHE_DIR` to move it, or delete the directory to free the
space.

# MERFISH

The MERFISH table has one row per cell: its original position, its position
aligned to the Visium section (`STalign_x`, `STalign_y`), the probability that
STalign matched it to the Visium section (`Pmatch`), and the counts of the
gene panel. We keep the cells matched with high confidence and rotate the
section by 90 degrees to the usual orientation.

```{r merfish}
merfish <- read.csv(files$merfish, row.names = 1)
dim(merfish)
names(merfish)[1:8]
matched <- merfish$Pmatch > 0.95
merfishCounts <- t(as.matrix(merfish[matched, -(1:5)]))
merfishXY <- cbind(x = merfish$STalign_x[matched],
                   y = -merfish$STalign_y[matched])
rownames(merfishXY) <- colnames(merfishCounts)
dim(merfishCounts)
```

# Visium

The Visium counts come as a Matrix Market file and the spot positions, in
pixels of the full-resolution image, as a table. We convert the positions to
the pixels of the high-resolution image, the frame of the STalign alignment,
and apply the same rotation.

```{r visium}
visiumDir <- tempfile("visium-")
untar(files$visium_counts, exdir = visiumDir)
untar(files$visium_spatial, exdir = visiumDir,
      files = c("spatial/tissue_positions_list.csv",
                "spatial/scalefactors_json.json"))
mtx <- file.path(visiumDir, "filtered_feature_bc_matrix")
visiumCounts <- methods::as(Matrix::readMM(file.path(mtx, "matrix.mtx.gz")),
                            "CsparseMatrix")
features <- read.delim(file.path(mtx, "features.tsv.gz"), header = FALSE)
barcodes <- readLines(file.path(mtx, "barcodes.tsv.gz"))
dimnames(visiumCounts) <- list(make.unique(features$V2), barcodes)

# columns: barcode, in_tissue, array_row, array_col, pixel row, pixel column
positions <- read.csv(file.path(visiumDir, "spatial", "tissue_positions_list.csv"),
                      header = FALSE, row.names = 1)
scale <- jsonlite::fromJSON(file.path(visiumDir, "spatial",
                                      "scalefactors_json.json"))$tissue_hires_scalef
visiumXY <- cbind(x = positions[barcodes, 5], y = -positions[barcodes, 4]) * scale
rownames(visiumXY) <- barcodes
dim(visiumCounts)
```

# Normalize and rasterize both technologies together

Only the genes measured by both technologies can be compared. To control for
the different detection efficiencies, each cell or spot is normalized by its
total counts over these genes, to counts per million (CPM).

```{r normalize}
genes <- intersect(rownames(visiumCounts), rownames(merfishCounts))
length(genes)
cpm <- function(m) Matrix::t(Matrix::t(m) / Matrix::colSums(m)) * 1e6
merfishSpe <- SpatialExperiment(
  assays = list(CPM = cpm(Matrix::Matrix(merfishCounts[genes, ], sparse = TRUE))),
  spatialCoords = merfishXY)
visiumSpe <- SpatialExperiment(assays = list(CPM = cpm(visiumCounts[genes, ])),
                               spatialCoords = visiumXY)
```

Cells and spots are then aggregated, in **one** call, onto a shared grid of
hexagonal pixels with `fun = "mean"`. The resolution, 20 pixels of the
high-resolution image, is about 80 µm, a little more than a Visium spot (55 µm
across, 100 µm apart), so that a pixel holds about one spot and many cells. The
assay `pixelval` holds the mean CPM of each pixel; for the correlation, we also
compute its logarithm, as in the published analysis.

```{r rasterize}
rast <- SEraster::rasterizeGeneExpression(
  list(MERFISH = merfishSpe, Visium = visiumSpe),
  assay_name = "CPM", resolution = 20, fun = "mean", square = FALSE)
for (s in names(rast)) {
  assay(rast[[s]], "lognorm") <- log10(assay(rast[[s]], "pixelval") + 1)
}
sapply(rast, ncol)
sapply(rast, function(r) median(r$num_cell))
```

Genes that are rarely detected cannot be compared, and a gene with no
expression in one technology cannot be tested. We keep the genes detected in
more than 1% of the pixels of both technologies:

```{r filter}
detected <- sapply(rast, function(r) Matrix::rowMeans(assay(r, "pixelval") > 0))
analysed <- rownames(detected)[detected[, "MERFISH"] > 0.01 &
                               detected[, "Visium"] > 0.01]
length(analysed)
```

To look at the pixels shared by both technologies, we draw each pixel as a
point at its centre (`SEraster::plotRaster()` draws the hexagons, which is
better suited to fewer, larger pixels). For example, Slc17a6 has a similar
pattern in both technologies:

```{r example-gene, fig.height = 3, fig.alt = "Maps of the log-transformed expression of Slc17a6 at the shared pixels, measured by MERFISH and by Visium: both show the same regions of high expression."}
shared <- intersect(colnames(rast$MERFISH), colnames(rast$Visium))
plotPixels <- function(spe, feature, title, assay = "lognorm") {
  values <- data.frame(spatialCoords(spe),
                       value = as.numeric(assay(spe, assay)[feature, ]))
  ggplot(values, aes(x, y, colour = value)) +
    geom_point(size = 0.4, shape = 16) +
    coord_equal() +
    scale_colour_viridis_c(name = NULL) +
    labs(title = title) +
    theme_void()
}
plotPixels(rast$MERFISH[, shared], "Slc17a6", "Slc17a6, MERFISH") +
  plotPixels(rast$Visium[, shared], "Slc17a6", "Slc17a6, Visium")
```

# Spatially variable genes

We expect genes with a spatial pattern in both technologies to be positively
correlated, and genes without a pattern not to be. The published analysis
classified the genes with Moran's I, as implemented in the GitHub package
[MERINGUE](https://github.com/JEFworks-Lab/MERINGUE):

```{r meringue, eval = FALSE}
# Not run: needs remotes::install_github("JEFworks-Lab/MERINGUE")
moransI <- function(mat, coords, filterDist) {
  w <- MERINGUE::getSpatialNeighbors(coords, filterDist = filterDist)
  I <- MERINGUE::getSpatialPatterns(mat, w)
  MERINGUE::filterSpatialPatterns(mat = mat, I = I, w = w, adjustPv = TRUE,
                                  alpha = 0.05, minPercentCells = 0.01)
}
# MERFISH: log10(CPM + 1) of the pixels; Visium: CPM of the spots
merfishI <- moransI(assay(rast$MERFISH, "lognorm")[analysed, ],
                    spatialCoords(rast$MERFISH), filterDist = 21)
visiumI <- moransI(assay(visiumSpe, "CPM"), spatialCoords(visiumSpe),
                   filterDist = 25)
svgGenes <- intersect(rownames(merfishI), rownames(visiumI))
```

STcompare ships the resulting list, so that this tutorial needs no GitHub-only
dependencies: the 230 of the `r length(analysed)` genes that are spatially
variable in both technologies.

```{r svg}
svgGenes <- scan(system.file("extdata", "vignette-brain-svg-genes.txt",
                             package = "STcompare"),
                 what = "", comment.char = "#", quiet = TRUE)
length(svgGenes)
nonSvgGenes <- setdiff(analysed, svgGenes)
length(nonSvgGenes)
```

# Spatial correlation of genes across technologies

We run the spatial correlation test on the log-transformed values. As in the
other case study, `nPermutations = 1000` keeps the run time short; see
[Parameters, performance and reproducibility](https://jef.works/STcompare/articles/parameters-performance-reproducibility.html)
for the trade-off.

```{r compare}
res <- compareSpatial(rast, assay = "lognorm", genes = analysed,
                      tests = "correlation", nPermutations = 1000,
                      nThreads = threads)
res
```

`compareSpatial()` does not test the `r sum(res$status == "skipped")` genes
that are detected in fewer than \(\sqrt{N}\) of the \(N\) =
`r res$nPixels[1]` shared pixels of a technology (here
`r ceiling(sqrt(res$nPixels[1]))` pixels; `minDetected`), most of them rarely
detected by Visium: the correlation of such genes is decided by a few pixels,
and their p-values would be too small. The message about `deltaGridEdge` says
that for a few genes, most surrogates chose the smallest (0.01) or the largest
smoothing bandwidth of the grid. Their p-values are still valid, but the
surrogates may match their spatial autocorrelation less well; a wider grid
(`delta`) would test that.

```{r svg-summary}
tab <- as.data.frame(res)
tab$svg <- ifelse(tab$gene %in% svgGenes, "spatially variable",
                  "not spatially variable")
table(tab$svg, tab$status)
tab <- tab[tab$status == "ok", ]
tab$significant <- tab$padj < 0.05
table(tab$svg, ifelse(tab$significant, ifelse(tab$r > 0, "positive", "negative"),
                      "not significant"))
```

`r sum(tab$significant & tab$r > 0 & tab$svg == "spatially variable")` of
`r sum(tab$svg == "spatially variable")`
(`r round(100 * mean(tab$significant[tab$svg == "spatially variable"] & tab$r[tab$svg == "spatially variable"] > 0))`%)
tested spatially variable genes are significantly positively correlated between
MERFISH and Visium, and
`r if (any(tab$significant & tab$r < 0)) paste(sum(tab$significant & tab$r < 0), "are") else "none is"`
significantly negatively correlated.
`r sum(!tab$significant & tab$svg == "not spatially variable")` of
`r sum(tab$svg == "not spatially variable")`
(`r round(100 * mean(!tab$significant[tab$svg == "not spatially variable"]))`%)
tested genes without a spatial pattern are not significantly correlated, as
expected.
The correspondence of the patterns is lower than between two MERFISH replicates
(93% of spatially variable genes in the paper), most likely because of the
different sensitivities of the two technologies and the different tissue
preservation (fresh frozen for MERFISH, FFPE for Visium) rather than a poor
alignment.

Two examples: the spatially variable gene with the strongest correlation, and
the gene without a spatial pattern with the lowest correlation among those that
are not significant. For each, the maps in both technologies and the values at
the shared pixels:

```{r gene-examples, fig.height = 3, fig.alt = "For two genes, maps of their log-transformed expression measured by MERFISH and by Visium, and a scatter plot of the two at each shared pixel: the first gene has matching patterns, the second does not."}
svgPositive <- tab[tab$svg == "spatially variable" & tab$significant & tab$r > 0, ]
geneSig <- svgPositive$gene[which.max(svgPositive$r)]
nonSvgNotSig <- tab[tab$svg == "not spatially variable" & !tab$significant, ]
geneNotSig <- nonSvgNotSig$gene[which.min(nonSvgNotSig$r)]

panels <- function(gene) {
  plotPixels(rast$MERFISH[, shared], gene, "MERFISH") +
    plotPixels(rast$Visium[, shared], gene, "Visium") +
    (plotCorrelationGeneExp(rast, res, gene, assayName = "lognorm") +
       theme(plot.title = element_text(size = 9))) +
    plot_layout(widths = c(1, 1, 1.3))
}
panels(geneSig)
panels(geneNotSig)
```

# Similarity of the levels

The similarity asks whether the two technologies measure similar levels at
the same pixels. It compares fold changes, so we compute it on the CPM
(`pixelval`), a linear scale; on log-transformed values, the ratio of two
values is not a fold change. The similarity test needs no permutations.

```{r similarity}
sim <- compareSpatial(rast, assay = "pixelval", genes = analysed,
                      tests = "similarity", verbose = FALSE)
tab$similarity <- sim[tab$gene, "similarity"]
summary(tab$similarity)
```

```{r similarity-plot, fig.height = 3.5, warning = FALSE, fig.alt = "Similarity against correlation for the significantly positively correlated spatially variable genes: most similarities are low."}
ggplot(subset(tab, svg == "spatially variable" & significant),
       aes(r, similarity)) +
  geom_point(alpha = 0.5, size = 1) +
  labs(x = "Correlation", y = "Similarity",
       title = "Significantly positively correlated spatially variable genes") +
  theme_classic()
```

Even for genes whose patterns agree, the levels at matched pixels often differ
more than two-fold (the median similarity of these genes is
`r round(median(tab$similarity[tab$svg == "spatially variable" & tab$significant], na.rm = TRUE), 2)`):
the two technologies capture each gene with a different efficiency, so the
fraction of a pixel's counts that a gene gets differs between them even after
normalization. Correlation, which ignores the levels,
is the appropriate test to compare technologies; similarity is more useful for
samples assayed in the same way.

# Cell types

While sensitivity to individual genes differs between technologies, the
underlying cell types should be the same. In the STalign paper (Clifton et
al. 2023), the MERFISH cells were clustered into cell types and the Visium
spots deconvolved into proportions of cell types, and the two sets of cell
types were matched by their transcriptional similarity:

```{r celltypes}
typeSimilarity <- as.matrix(read.csv(files$type_similarity, row.names = 1,
                                     check.names = FALSE))
colnames(typeSimilarity) <- rownames(typeSimilarity)
merfishTypes <- read.csv(files$merfish_types, row.names = 1)
visiumTypes <- as.matrix(read.csv(files$visium_types, row.names = 1,
                                  check.names = FALSE))
colnames(visiumTypes) <- rownames(typeSimilarity)

# cell types whose best transcriptional match is mutual
best <- rownames(typeSimilarity)[
  diag(typeSimilarity) == apply(typeSimilarity, 1, max) &
  diag(typeSimilarity) == apply(typeSimilarity, 2, max)]
best
```

To compare where the cell types are, each cell type becomes a row of an assay:
a one-hot encoding of the MERFISH cells' labels, and the deconvolved
proportions of the Visium spots. Rasterized together with `fun = "mean"`, the
pixels hold the proportion of each cell type.

```{r celltype-rasterize}
labelled <- intersect(colnames(merfishSpe), rownames(merfishTypes))
labels <- factor(merfishTypes[labelled, "x"], levels = rownames(typeSimilarity))
oneHot <- t(sapply(levels(labels), function(type) as.numeric(labels == type)))
colnames(oneHot) <- labelled
rastTypes <- SEraster::rasterizeGeneExpression(
  list(MERFISH = SpatialExperiment(assays = list(proportion = oneHot),
                                   spatialCoords = merfishXY[labelled, ]),
       Visium = SpatialExperiment(assays = list(proportion = t(visiumTypes)),
                                  spatialCoords = visiumXY[rownames(visiumTypes), ])),
  assay_name = "proportion", resolution = 20, fun = "mean", square = FALSE)
```

```{r celltype-maps, fig.height = 4.5, fig.alt = "Maps of the proportions of three matched cell types at the shared pixels, measured by MERFISH (top) and by Visium (bottom)."}
sharedTypes <- intersect(colnames(rastTypes$MERFISH),
                         colnames(rastTypes$Visium))
typeMaps <- lapply(names(rastTypes), function(s) {
  lapply(head(best, 3), function(type) {
    plotPixels(rastTypes[[s]][, sharedTypes], type, paste0(type, ", ", s),
               assay = "pixelval")
  })
})
wrap_plots(unlist(typeMaps, recursive = FALSE), nrow = 2)
```

The proportions are on a linear scale, so both tests apply:

```{r celltype-compare}
resTypes <- compareSpatial(rastTypes, nPermutations = 1000, nThreads = threads)
tabTypes <- as.data.frame(resTypes)
tabTypes$matched <- tabTypes$gene %in% best
tabTypes[order(-tabTypes$matched, tabTypes$p),
         c("gene", "matched", "r", "padj", "nPermutations", "similarity")]
```

```{r celltype-count, include = FALSE}
matchedPositive <- sum(tabTypes$matched & tabTypes$padj < 0.05 & tabTypes$r > 0,
                       na.rm = TRUE)
skippedTypes <- tabTypes[tabTypes$matched & tabTypes$status == "skipped", ]
one <- nrow(skippedTypes) == 1L
skippedText <- if (nrow(skippedTypes)) {
  sprintf(paste("%s %s in fewer than %d of the shared pixels of a technology, so",
                "%s not tested (`minDetected`), although %s %s high (r = %s):",
                "with so few pixels, the surrogates cannot give a reliable",
                "p-value."),
          if (one) "The other," else "The others,",
          paste0(paste(skippedTypes$gene, collapse = " and "), if (one) ", is" else ", are"),
          ceiling(sqrt(resTypes$nPixels[1])), if (one) "it is" else "they are",
          if (one) "its correlation" else "their correlations", if (one) "is" else "are",
          paste(round(skippedTypes$r, 2), collapse = " and "))
} else ""
```

`r if (matchedPositive == length(best)) paste("All", length(best)) else paste(matchedPositive, "of the", length(best))`
cell types with a mutual transcriptional match are significantly positively
correlated between the technologies: their spatial patterns correspond, even
though the patterns of many individual genes do not. `r skippedText`

# Session information

```{r session}
sessionInfo()
```
